---
title: "Multi Ancestry Analysis"
output: html_document
---

<!-- TODO
- `input_dir` (line ~9): `drive/MyDrive/statgen_workshop/data/workshop7/` is an incomplete Colab path (missing `/content/` prefix) — update to real data path
- `analysis_dir` is already set to `$(pwd)/07_analysis/` which is portable — no change needed
- `cp /content/drive/MyDrive/statgen_workshop/tools/07_multi_ancestry_analysis_R_libs.tgz ./` (line ~39): Update path to the pre-built library tarball, or install packages directly
- `system(paste0("/content/MsCAVIAR//MsCAVIAR ..."))` (line ~344): Update `/content/MsCAVIAR/` to real path where MsCAVIAR binary is installed
- Commented-out `# python /content/PRScsx/PRScsx.py` (line ~407): If activating this, update to real path for PRScsx.py
-->

# Set up

```{bash}
export input_dir="drive/MyDrive/statgen_workshop/data/workshop7/"
export analysis_dir="$(pwd)/07_analysis/"
```

Load R magic

# PART 1. Multi-population Fine-mapping
## A single-hit region on chr 12

### mJAM

Step 1. Packages

Install dependencies for the hJAM R libary

```{bash}
apt-get install libgsl-dev libmpfr-dev
```

We need to install the hJAM and tidyverse R libaries. Here's the code to install them:


```
devtools::install_github("USCbiostats/hJAM")
if(!"tidyverse"%in%installed.packages()){install.packages('tidyverse')}
```

In the interest of time, we will add the pre-installed packages to R path so we don't have to wait for the R package installations.

```{bash}
cp /content/drive/MyDrive/statgen_workshop/tools/07_multi_ancestry_analysis_R_libs.tgz ./ && \
   tar -zxf 07_multi_ancestry_analysis_R_libs.tgz && ls usr/local/lib/R/site-library/
```

Add to R library Path

```{r}
.libPaths("usr/local/lib/R/site-library/")
.libPaths()
```

load R libraries

```{r}
library(hJAM) ## for mJAM_Forward, mJAM_SuSiE
library(tidyverse) ## for data cleaning and visualization
library(glue)
```

Visualize marginal results

```{r}
meta_sumstat <- read.table(glue(input_dir, "MultiPopulation_FineMapping/12_109194870_110794870_meta_summary_stat.txt"), header = T)
```

Meta P.value

```{r}
meta_sumstat %>%
  ggplot(aes(x = POS, y = -log10(as.numeric(P.value))))+
  geom_point(size = 1)+
  geom_hline(yintercept = -log10(5e-8), linetype="dashed", color = "red")+
  labs(x = "Position", y = "-log10(meta.p.value)",
       title = "Marginal signficance of meta-analyzed summary statistics")
```

Population-specific P.value

```{r}
meta_sumstat %>%
  select(c(POS, EUR.P.value, AFR.P.value, HIS.P.value, EAS.P.value)) %>%
  mutate(EUR.P.value = as.numeric(EUR.P.value),
         AFR.P.value = as.numeric(AFR.P.value),
         HIS.P.value = as.numeric(HIS.P.value),
         EAS.P.value = as.numeric(EAS.P.value)) %>%
  pivot_longer(cols = EUR.P.value:EAS.P.value,
               names_to = "pop",
               values_to = "P.value") %>%
  ggplot(aes(x = POS, y = -log10(as.numeric(P.value))))+
  geom_point(size = 1)+
  geom_hline(yintercept = -log10(5e-8), linetype="dashed", color = "red")+
  facet_wrap(vars(pop), nrow = 2, scales = "free_y")+
  labs(x = "Position", y = "-log10(p.value)",
       title = "Marginal signficance by population")
```

Run various fine-mapping approaches

#### mJAM-Forward

```{r}
Marg_Result <- read.table(glue(input_dir, "MultiPopulation_FineMapping/1_mJAM/data/Marg_results.txt"), header = T)
MAF_Result <- read.table(glue(input_dir, "MultiPopulation_FineMapping/1_mJAM/data/MAF_results.txt"), header = T)
EUR_ONCO_dosage <- read.table(glue(input_dir, "MultiPopulation_FineMapping/1_mJAM/data/EUR_ONCO_dosages.txt"), header = T, check.names = F)
AFR_ONCO_dosage <- read.table(glue(input_dir, "MultiPopulation_FineMapping/1_mJAM/data/AFR_ONCO_dosages.txt"), header = T, check.names = F)
HIS_ONCO_dosage <- read.table(glue(input_dir, "MultiPopulation_FineMapping/1_mJAM/data/HIS_ONCO_dosages.txt"), header = T, check.names = F)
EAS_ONCO_dosage <- read.table(glue(input_dir, "MultiPopulation_FineMapping/1_mJAM/data/EAS_ONCO_dosages.txt"), header = T, check.names = F)

## sample size of discovery GWAS (as in summary statistics)
N_cases_EUR <- 122188; N_ctrl_EUR <- 604640
N_cases_AFR <- 19391; N_ctrl_AFR <- 61608
N_cases_HIS <- 3931; N_ctrl_HIS <- 26405
N_cases_EAS <- 10809; N_ctrl_EAS <- 95790
N_GWAS_EUR <- 4/(1/N_cases_EUR + 1/N_ctrl_EUR)
N_GWAS_AFR <- 4/(1/N_cases_AFR + 1/N_ctrl_AFR)
N_GWAS_HIS <- 4/(1/N_cases_HIS + 1/N_ctrl_HIS)
N_GWAS_EAS <- 4/(1/N_cases_EAS + 1/N_ctrl_EAS)
```

fit mJAM with forward selection

```{r}
mJAM_Forward_fit <- hJAM::mJAM_Forward(N_GWAS = c(N_GWAS_EUR, N_GWAS_AFR, N_GWAS_HIS, N_GWAS_EAS),
                                       X_ref = list(EUR_ONCO_dosage, AFR_ONCO_dosage, HIS_ONCO_dosage, EAS_ONCO_dosage),
                                       Marg_Result = Marg_Result,
                                       EAF_Result = MAF_Result,
                                       condp_cut = 5e-08,
                                       within_pop_threshold = 0.20,
                                       across_pop_threshold = 0.10)
```

index SNP's selected by mJAM_Forward

```{r}
mJAM_Forward_fit$index
```

utility function for plotting

```{r}
plot_fine_mapping_res <- function(by.pop = FALSE,
                                  selected_index,
                                  cs_snp){

  if(by.pop){
    cs_snp_by_pop = cs_snp %>%
      select(c(index_SNP, CS_SNP, is_index, POS, EUR.P.value, AFR.P.value, HIS.P.value, EAS.P.value)) %>%
      mutate(EUR.P.value = as.numeric(EUR.P.value),
             AFR.P.value = as.numeric(AFR.P.value),
             HIS.P.value = as.numeric(HIS.P.value),
             EAS.P.value = as.numeric(EAS.P.value)) %>%
      pivot_longer(cols = EUR.P.value:EAS.P.value,
                   names_to = "pop",
                   values_to = "P.value")
    meta_sumstat %>%
      select(c(MarkerName, POS, EUR.P.value, AFR.P.value, HIS.P.value, EAS.P.value)) %>%
      mutate(EUR.P.value = as.numeric(EUR.P.value),
             AFR.P.value = as.numeric(AFR.P.value),
             HIS.P.value = as.numeric(HIS.P.value),
             EAS.P.value = as.numeric(EAS.P.value)) %>%
      pivot_longer(cols = EUR.P.value:EAS.P.value,
                   names_to = "pop",
                   values_to = "P.value") %>%
      ggplot(aes(x = POS, y = -log10(P.value)))+
      geom_point(color="grey", size  = 1.5)+
      geom_point(aes(color = index_SNP, group = index_SNP), data = cs_snp_by_pop, size  = 1)+
      geom_point(data = filter(cs_snp_by_pop, is_index == T),
                 colour="red", shape=1, size=1.5, stroke=1)+
      geom_hline(yintercept = -log10(5e-8), linetype="dashed", color = "red")+
      facet_wrap(vars(pop), nrow = 2, scales = "free_y")+
      labs(x = "Position", y = "-log10(p.value)",
           title = "Index SNPs and credible sets (by population)")
  }else{
    meta_sumstat %>%
      ggplot(aes(x = POS, y = -log10(as.numeric(P.value))))+
      geom_point(color="grey", size  = 1.5)+
      geom_point(aes(color = index_SNP, group = index_SNP), data = cs_snp, size  = 1)+
      geom_point(data = filter(cs_snp, is_index == T),
                 colour="red", shape=1, size=1.5, stroke=1)+
      geom_hline(yintercept = -log10(5e-8), linetype="dashed", color = "red")+
      labs(x = "Position", y = "-log10(p.value)",
           title = "Index SNPs and credible sets")
  }

}
```

pull variants in mJAM-Forward credible sets

```{r}
selected_index_forward <- mJAM_Forward_fit$index$SNP
cs_snp_forward <- mJAM_Forward_fit$cs %>%
  filter(CS_in == TRUE) %>%
  left_join(meta_sumstat, by = c("CS_SNP"="MarkerName")) %>%
  mutate(is_index = (index_SNP == CS_SNP)) %>%
  left_join(tibble(index_SNP = selected_index_forward,
                   oldname = meta_sumstat$MarkerName_old[match(selected_index_forward,meta_sumstat$MarkerName)],
                   order = 1:length(selected_index_forward)), by = "index_SNP") %>%
  mutate(index_SNP_legend = paste0("(#",order,")",oldname)) %>%
  select(-index_SNP) %>%
  rename(index_SNP = index_SNP_legend)
```

mJAM-Forward index SNPs and credible sets (meta P values)

```{r}
plot_fine_mapping_res(by.pop = FALSE,
                      selected_index = selected_index_forward,
                      cs_snp = cs_snp_forward)
```

mJAM-Forward index SNPs and credible sets (population-specific P values)

```{r}
plot_fine_mapping_res(by.pop = TRUE,
                      selected_index = selected_index_forward,
                      cs_snp = cs_snp_forward)
```

#### mJAM-SuSiE

fit mJAM with Sum of Single Effect (Bayesian feature selection)

```{r}
mJAM_SuSiE_fit <- hJAM::mJAM_SuSiE(Marg_Result = Marg_Result,
                                   EAF_Result = MAF_Result,
                                   N_GWAS = c(N_GWAS_EUR, N_GWAS_AFR, N_GWAS_HIS, N_GWAS_EAS),
                                   X_ref = list(EUR_ONCO_dosage, AFR_ONCO_dosage, HIS_ONCO_dosage, EAS_ONCO_dosage))


mJAM_SuSiE_CS_0.95_Output <- hJAM::mJAM_SuSiE_get_cs(mJAM_SuSiE_fit,0.95)
```

pull mJAM-SuSiE index SNPs and credible sets

```{r}
selected_index_susie <- mJAM_SuSiE_CS_0.95_Output$index_SNP
cs_snp_susie <- mJAM_SuSiE_CS_0.95_Output %>%
  left_join(meta_sumstat, by = c("CS_SNP"="MarkerName")) %>%
  mutate(is_index = (index_SNP == CS_SNP))
```

mJAM-SuSiE index SNPs and credible sets (meta P values)

```{r}
plot_fine_mapping_res(by.pop = FALSE,
                      selected_index = selected_index_susie,
                      cs_snp = cs_snp_susie)
```

mJAM-SuSiE index SNPs and credible sets (population-specific P values)

```{r}
plot_fine_mapping_res(by.pop = TRUE,
                      selected_index = selected_index_susie,
                      cs_snp = cs_snp_susie)
```

## COJO

Step 1: installation: https://yanglab.westlake.edu.cn/software/gcta/#Download

```{bash}
wget https://yanglab.westlake.edu.cn/software/gcta/bin/gcta-1.94.1-linux-kernel-3-x86_64.zip && unzip gcta-1.94.1-linux-kernel-3-x86_64.zip
```

```{python}
ls gcta-1.94.1-linux-kernel-3-x86_64/
```

Step 2: Running COJO using command line

```{bash}
mkdir -p ${analysis_dir}/2_COJO/results/

./gcta-1.94.1-linux-kernel-3-x86_64/gcta-1.94.1 \
--bfile ${input_dir}/MultiPopulation_FineMapping/2_COJO/data/12_109194870_110794870 \
--cojo-file ${input_dir}/MultiPopulation_FineMapping/2_COJO/data/12_109194870_110794870.ma \
--cojo-slct --cojo-p 5e-08 \
--out ${analysis_dir}/2_COJO/results/AllPopRes
```

Read the results in R

```{r}
COJO_res <- read.table(glue(analysis_dir, "2_COJO/results/AllPopRes", ".jma.cojo"),
                       header = T, sep = '\t')
```

note that COJO only provides index SNP selection

```{r}
COJO_res
```

plot COJO index SNPs

```{r}
meta_sumstat %>%
  ggplot(aes(x = POS, y = -log10(as.numeric(P.value))))+
  geom_point(color = "grey")+
  geom_point(data = meta_sumstat[meta_sumstat$MarkerName %in% COJO_res$SNP,],
             color = 'black')+
  geom_point(data = meta_sumstat[meta_sumstat$MarkerName %in% COJO_res$SNP,],
             colour="red", shape=1, size=2.5, stroke=1.5)+
  geom_text(
    data = meta_sumstat[match(COJO_res$SNP, meta_sumstat$MarkerName),],
    label= COJO_res$SNP,
    nudge_x = 0.25, nudge_y = 0.3, check_overlap = T, size = 2.5
  )+
  labs(x = "Position", y = "-log10(P.value)",
       title = "COJO index SNPs")
```

## MsCAVIAR

Step1: installation: https://github.com/nlapier2/MsCAVIAR

```{bash}
git clone https://github.com/nlapier2/MsCAVIAR.git && cd MsCAVIAR/ && make
```

Step 2: prepare files for MsCAVIAR

```{r}
msCAVIAR_Dir = glue(input_dir, "/MultiPopulation_FineMapping/3_msCAVIAR/data/")
msCAVIAR_output_Dir = glue(analysis_dir, "3_msCAVIAR/results/")
dir.create(msCAVIAR_output_Dir, recursive=TRUE)
```

LD and zscores for each population

```{r}
all.files = list.files(path = msCAVIAR_Dir)
ld.files = all.files[ endsWith(all.files, paste0("_LD.txt")) ]
write.table(paste0(msCAVIAR_Dir, ld.files), paste0(msCAVIAR_output_Dir, "ld.files.txt"), quote = F, col.names = F, row.names = F)
zscore.files = all.files[ endsWith(all.files, paste0("_zscore.txt")) ]
write.table(paste0(msCAVIAR_Dir, zscore.files), paste0(msCAVIAR_output_Dir, "zscore.files.txt"), quote = F,  col.names = F, row.names = F)

N_Sample = c(N_GWAS_EUR, N_GWAS_AFR, N_GWAS_HIS, N_GWAS_EAS)
```

Step 3: run MsCAVIAR using command line tool or in R

```{r}
system(paste0("/content/MsCAVIAR//MsCAVIAR -l ",
              paste0( msCAVIAR_output_Dir, "ld.files.txt"),
              " -z ", paste0( msCAVIAR_output_Dir, "zscore.files.txt"),
              " -n ", paste(round(N_Sample), collapse = ","),
              " -o ", paste0(msCAVIAR_output_Dir, "output_12_109194870_110794870"),
              " -c ", 1), intern = T) ## allow max causal = 1
```

Step 4: read MsCAVIAR results

```{r}
mscaviar_cs_snps = read.table(paste0(msCAVIAR_output_Dir, "output_12_109194870_110794870_set.txt"), sep = "\t", header=F)
mscaviar_pip = read.table(paste0(msCAVIAR_output_Dir, "output_12_109194870_110794870_post.txt"), sep = "\t", header=T)
mscaviar_index = mscaviar_pip %>%
  left_join(meta_sumstat[,c("MarkerName", "MarkerName_old")], by = c("SNP_ID" = "MarkerName")) %>%
  arrange(desc(Prob_in_pCausalSet)) %>% head(1) %>% pull(MarkerName_old)
```

Step 5: plot MsCAVIAR results

```{r}
meta_sumstat %>%
  ggplot(aes(x = POS, y = -log10(as.numeric(P.value))))+
  geom_point(color = "grey")+
  geom_point(data = filter(meta_sumstat, MarkerName %in% mscaviar_cs_snps$V1) ,
             color = "blue")+
  geom_point(data = filter(meta_sumstat,  MarkerName_old %in% mscaviar_index),
             colour="red", shape=1, size=2.5, stroke=1.5)+
  geom_label(
    data = filter(meta_sumstat,  MarkerName_old %in% mscaviar_index),
    label= mscaviar_index,
    nudge_x = 0.25, nudge_y = 0.9, size = 4
  )+
  geom_hline(yintercept = -log10(5e-8), linetype="dashed", color = "red")+
  labs(x = "Position", y = "-log10(meta P.value)", title = "MsCAVIAR index SNPs and credible sets")
```

# Genomewide PRS

## Multi-population GW PRS with PRS-CSx

Step1: PRS-CSx installation https://github.com/getian107/PRScsx

Please note that the h5py and scipy libraries might be needed in your own runtime (they are pre-installed in the colab runtime).

```{bash}
git clone https://github.com/getian107/PRScsx.git
```

Step2: take a look at population-specific PCa summary statistics on chr22

```{r}
library(glue)
library(tidyverse)
sumstat_EUR <- read.table(glue(input_dir, "Genomewide_PRS/data/summaryStats_hm3_chr22.EUR.txt"), header = T)
head(sumstat_EUR)
```

Step3: Run PRS-CSx script using command line - this step takes more than an hour to run on the free version of colab. So we provide the commented out code here but will be using pre-run results for the rest of the tutorial

```{bash}
# mkdir -p ${analysis_dir}/Genomewide_PRS/results/

# python /content/PRScsx/PRScsx.py \
#  --ref_dir=${input_dir}/Genomewide_PRS/ldRefPanel/ \
#  --bim_prefix=${input_dir}/Genomewide_PRS/data/H3.hm3.chr22.rsid \
#  --sst_file=${input_dir}/Genomewide_PRS/data/summaryStats_hm3_chr22.AFR.txt,${input_dir}/Genomewide_PRS/data/summaryStats_hm3_chr22.EUR.txt,${input_dir}/Genomewide_PRS/data/summaryStats_hm3_chr22.EAS.txt,${input_dir}/Genomewide_PRS/data/summaryStats_hm3_chr22.HIS.txt \
#  --n_gwas=58995,406587,38851,13686 \
#  --pop=AFR,EUR,EAS,AMR \
#  --meta=TRUE \
#  --chrom=22 \
#  --out_dir=${analysis_dir}/Genomewide_PRS/results/ \
#  --out_name=test_chr22
```

Step4: read results
Note that we are reading in the pre-generated results here, please change the file path if you have run the PRS-CSx script

```{r}
prscsx_res <- read.table(glue(input_dir, "Genomewide_PRS/results/test_chr22_META_pst_eff_a1_b0.5_phiauto_chr22.txt"), header = F)
colnames(prscsx_res) <- c("chr", "rsid", "pos", "effect.allele", "other.allele", "prscsx_weight")
head(prscsx_res)
```

Step5: visualize weights before vs after PRS-CSx

```{r}
meta_sumstat_chr22 <- read.table(glue(input_dir,"Genomewide_PRS/data/meta_summaryStats_hm3_chr22.txt"), header = T)

meta_sumstat_chr22 %>%
  left_join(prscsx_res, by = c("chr","pos","rsid")) %>%
  mutate(flip_prscsx_weight = (Allele1 == other.allele),
         prscsx_weight_flipped = ifelse(flip_prscsx_weight, (-1)*prscsx_weight, prscsx_weight)) %>%
  ggplot(aes(x = pos))+
  geom_point(aes(y = abs(Effect)), alpha = 0.5, size=1, color = "grey")+
  geom_point(aes(y = abs(prscsx_weight_flipped)), alpha = 0.5, size=1, color = "blue")+
  labs(title = "PRS-CSx weights (blue) vs Marginal effect sizes (grey)",
       y = "weight", x = "position on chr22")
```

# PRS Evaluation

### Import Libraries

Install pROC

```{r}
install.packages("pROC")
```

```{r}
library(dplyr)
library(ggplot2)
library(tidyr)
library(purrr) # purrr function
library(pROC) # roc for glm models
library(stringr)
```

### Define Global Vars

```{r}
data.loc = glue(input_dir, "PRS_Evaluation/")
data.loc
```

### Define Functions

```{r}

# pull the weights needed for the PRS calculation for each variant from meta-data (pull info for dosage)
row_pull = function(in.chr, in.pos, in.ref, in.alt, wt.df.name){
  wt.df = get(wt.df.name)

    wt.df[wt.df$chr == in.chr & wt.df$pos == in.pos &
            ((wt.df$effect_allele == in.ref & wt.df$other_allele == in.alt) |
               (wt.df$other_allele == in.ref & wt.df$effect_allele == in.alt)), ]
}

# flip dosages in dosage file to reflect added risk
correct_dosage = function(in.dosage){
  chrpos.tmp = cur_column()
  # compare risk allele against alt and flip if mismatch (check what alt refers to: )
  if (meta.data.wts[meta.data.wts$chr_pos_alt_ref == chrpos.tmp, "wt_flip"][[1]] == 1){
    abs(in.dosage-2)
  } else {
    in.dosage
  }
}
```

### Import Data

```{r}

dta.pc.dosage =
  glue(data.loc, "dta_PC_PRS_dosage.rds") %>%
  readRDS()

dta.meta.data =
  glue(data.loc, "dta.PC.MetaData.rds") %>%
  readRDS()

# remove effect_weight_new
PRS.weights.data =
  glue(data.loc, "PRS.weights.rds") %>%
  readRDS()
```

look at all the data to get an idea of what it is

```{r}
dta.pc.dosage[1:10,1:30] %>% as.data.frame()
```

```{r}
dta.pc.dosage %>% group_by(eth, cancer) %>% tally()
```

```{r}
dta.meta.data %>% head() %>% as.data.frame()
```

```{r}
PRS.weights.data %>% head %>% as.data.frame()
```

### Prep Data

1) append the weights information (from literature) to our meta-data file

```{r}
meta.data.wts =
  dta.meta.data %>%
  # use our function to create a new column with weights from the weights file
  # this is basically merging based on chromosome, position, allele 1, and allele 2, it is a
  # custom function to handle multialleleic cases
  mutate(prs_info = pmap(.f=row_pull, list(CHR, POS, REF, ALT, "PRS.weights.data")),
         prs_info_n = prs_info %>% map_int(.f=function(x){nrow(x)})) %>%
  unnest_wider(prs_info, names_sep="_") %>%
  # filter out variants from our meta-data that we are not interested in
  filter(prs_info_n == 1) %>%

  # identify what variants need to have their dosage transformed to reflect added risk
  # for PRS calculation
  mutate(wt_flip = ifelse(prs_info_effect_allele != ALT, 1, 0))

meta.data.wts %>% head %>% as.data.frame()
```

2) use our wt_flip variable to recode the dosages as "risk" based on meta-data file

```{r}
dta.pc.dosage.risk =
  dta.pc.dosage %>%
  # select all covariates from individual-level dosage file, then select all
  # variants in our meta data file (variants to be used in the PRS calculation)
  select(-contains(":"), meta.data.wts$chr_pos_alt_ref) %>%
  # mutate across all dosages (chrpos), pass to flip function that returns transformed dosage vector
  mutate(across(contains(":"), correct_dosage))
```

### PRS Calculation

calculate unweighted PRS

create a new variable prs_uwt which is just a sum of all PRS risk alleles, across all variants

```{r}
dta.pc.dosage.risk$prs_uwt =
  dta.pc.dosage.risk %>%
  select(meta.data.wts$chr_pos_alt_ref) %>%
  as.matrix() %>%
  apply(., 1, sum)
```

calculate weighted PRS

Sum all risk alleles, after applying weights to each variant, done using matrix multiplication

```{r}
dta.pc.dosage.risk$prs_wtd =
  c(as.matrix(select(dta.pc.dosage.risk, meta.data.wts$chr_pos_alt_ref)) %*% meta.data.wts$prs_info_effect_weight)
```

### Descriptives of PRS

```{r}
dta.pc.dosage.risk %>%
  ggplot(aes(x=prs_wtd, color=as.factor(cancer)))+
  geom_density() +
  facet_wrap(~eth)
```

```{r}
dta.pc.dosage.risk %>%
  ggplot(aes(x=prs_wtd, fill=eth, color=eth))+
  geom_density(alpha=.5)
```

## Apply Standardization Techniques

## Ancestry-Residual Model for removal of "relatedness/ancestry" variation in PRS

```{r}
dta.pc.dosage.risk$mdl.resid =
  lm(data=dta.pc.dosage.risk,
     formula = as.formula("prs_wtd ~ V1 + V2 + V3 + V4 + V5 + V6 + V7 + V8 + V9 + V10"))$residuals
```

standardize new 'residualized PRS"

```{r}
prs_resid_stats =
  dta.pc.dosage.risk %>%
  # select controls
  filter(cancer == 0) %>%
  summarise(mean_prs_resid = mean(mdl.resid), sd_prs_resid = sd(mdl.resid))
```

```{r}
dta.pc.dosage.risk$prs_mdl_resid_std =
  (dta.pc.dosage.risk$mdl.resid - prs_resid_stats$mean_prs_resid)/prs_resid_stats$sd_prs_resid
```

Standardized within Race/Eth

```{r}
prs_eth_stats_eth =
  dta.pc.dosage.risk %>%
  # select controls
  filter(cancer == 0) %>%
  group_by(eth) %>%
  summarise(mean_prs_wts_eth = mean(prs_wtd), sd_prs_wtd_eth = sd(prs_wtd))
```

```{r}
prs_eth_stats_combined =
  dta.pc.dosage.risk %>%
  # select controls
  filter(cancer == 0) %>%
  summarise(mean_prs_wts_combined = mean(prs_wtd), sd_prs_wtd_combined = sd(prs_wtd))
```

```{r}
dta.pc.dosage.risk =
  dta.pc.dosage.risk %>%
  left_join(prs_eth_stats_eth) %>%
  mutate(prs_wtd_eth_std = (prs_wtd - mean_prs_wts_eth)/sd_prs_wtd_eth,
         prs_wtd_std = (prs_wtd - prs_eth_stats_combined$mean_prs_wts_combined)/
           prs_eth_stats_combined$sd_prs_wtd_combined)
```

## Compare All Types of PRS we estimated

compare prs distributions across race/eth, combined, among controls

```{r}
prs.compare.plt.data =
  rbind(dta.pc.dosage.risk %>% select(prs_wtd_std, prs_wtd_eth_std, prs_mdl_resid_std) %>%
          mutate(eth = "Combined"),
        dta.pc.dosage.risk %>% select(prs_wtd_std, prs_wtd_eth_std, prs_mdl_resid_std, eth)) %>%
  pivot_longer(starts_with("prs"), values_to = "PRS_Value", names_to = "PRS_type")
```

```{r}
prs.compare.plt.data %>%
  ggplot(aes(x=PRS_Value, color=PRS_type, fill=PRS_type))+
  geom_density(alpha=.4) +
  facet_wrap(~eth, nrow = 1)
```

```{r}
prs.compare.plt.data %>%
  ggplot(aes(x=PRS_Value, color=eth, fill=eth))+
  geom_density(alpha=.4) +
  facet_wrap(~PRS_type, nrow = 1)
```

## Regression Analysis

Simple PRS model with age and PC adjustment

```{r}
# define formula
fm = as.formula("cancer ~ prs_wtd_std + age_draw + V1 + V2 + V3 + V4 + V5 + V6")

# fit model
res = glm(fm, data = dta.pc.dosage.risk, binomial(link='logit'))
res
```

view coefficient results as data.frame

```{r}
res %>% summary() %>% pluck("coefficients") %>% as.data.frame()
```

check aic

```{r}
res$aic
```

check AUC

```{r}
res_pred_prob = predict(res, newdata = dta.pc.dosage.risk, type = "response")
roc(dta.pc.dosage.risk$cancer ~ res_pred_prob, plot = TRUE, print.auc = TRUE)
```

AUC is 0.7075, lets compare against a model with the covariates (which may be the ones actually producing a high auc)

fit model

```{r}
null.mdl.res = glm(as.formula("cancer ~ age_draw + V1 + V2 + V3 + V4 + V5 + V6"),
                   data = dta.pc.dosage.risk, binomial(link='logit'))

res_pred_prob_null_mdl = predict(null.mdl.res, newdata = dta.pc.dosage.risk, type = "response")
roc(dta.pc.dosage.risk$cancer ~ res_pred_prob_null_mdl, plot = TRUE, print.auc = TRUE)
```

the null model produces an AUC of 0.5832, versus including PRS which is .71, this is a considerable increase due to PRS.

## Compare a variety of models

prs terms to evaluate: prs_wtd_std, prs_wtd_eth_std, prs_mdl_resid_std

function to run regression and pull PRS term and other fit stats

```{r}
fit_prs_models = function(in.model, in.eth, in.df.name){
  dta.tmp = get(in.df.name)

  # filter data
  if (in.eth != "Combined"){
    dta.tmp =
      dta.tmp %>%
      filter(eth == in.eth)
  }

  # fit model:
  res = glm(as.formula(in.model), data = dta.tmp, binomial(link='logit'))

  # extract prs terms
  res.df =
    res %>% summary() %>% pluck("coefficients") %>% as.data.frame() %>%
    mutate(var = row.names(.)) %>%
    filter(grepl("prs_", var))

  # check aic
  res.df$aic = res$aic

  # check AUC
  res_pred_prob = predict(res, newdata = dta.pc.dosage.risk, type = "response")
  res.df$auc = roc(dta.pc.dosage.risk$cancer ~ res_pred_prob, plot = F, print.auc = F)$auc[[1]]

  return(res.df)
}
```

function to help build results/models data frame

```{r}
dup_over = function(in.df, in.varname, in.levels){
  in.df = in.df %>%
    mutate(tmp_in_varname = 1 %>%
             map(.f=function(x){data.frame(tmp_in_varname=in.levels)})) %>%
    unnest(tmp_in_varname)

  names(in.df)[names(in.df)=="tmp_in_varname"] = in.varname
  in.df
}
```

build results/model dataframe and run regressions

```{r}
all.fit.res =
  data.frame(eth = c(unique(dta.pc.dosage.risk$eth), "Combined")) %>%
  dup_over("PRS", c("prs_wtd_std", "prs_wtd_eth_std", "prs_mdl_resid_std")) %>%
  dup_over("model_type", c("cancer ~ prs",
                            "cancer ~ prs + age_draw",
                            "cancer ~ prs + age_draw + V1 + V2 + V3 + V4 + V5 + V6")) %>%
  mutate(model = str_replace(model_type, "prs", PRS),
         model_type = ifelse(grepl("V1", model_type), "cancer ~ prs + age_draw + PCs", model_type)) %>%

  # map analyses over fit parameters (run regressions)
  mutate(fit_results = pmap(.f=fit_prs_models, list(model, eth, "dta.pc.dosage.risk"))) %>%
  # unnest returned results
  unnest_wider(fit_results)
```

View PRS Beta Coefficents (exp(beta) for HRs)

```{r}
all.fit.res %>%
  mutate(beta_lci = Estimate-(1.96*`Std. Error`),
         beta_uci = Estimate+(1.96*`Std. Error`)) %>%

  ggplot(aes(x=Estimate, y=model_type, color=PRS, xmin=beta_lci, xmax=beta_uci, label=sprintf("%.2f", auc)))+
  geom_point(position = position_dodge(width=.5), size=2.5) +
  geom_linerange(position = position_dodge(width=.5), linewidth = 1.5) +
  geom_text(position = position_dodge(width=.5), x=.4, size=2.5)+
  xlim(.3,1.3)+
  facet_wrap(~eth) +
  theme_light()
```

