---
title: "Heritability and polygenic scores"
subtitle: "Statistical genetics practical in RStudio on Biowulf"
output:
  html_document:
    toc: true
    toc_float: false
    number_sections: true
    code_folding: show
    self_contained: true
    theme: readable
params:
  execute: true
  data_dir: "/data/DCEG_shared/statgen_workshop_2026/data/statgen_workshop_04Heritability_PRS/data"
  gcta_phenotype: ""
  output_dir: "results"
  threads: 1
  plink_memory_mb: 2048
  run_gcta: true
  run_ldsc: true
  run_sldsc: true
  run_prs: true
  tool_mode: "biowulf_modules"
  gcta_bin: "gcta64"
  plink_bin: "plink"
  plink2_bin: "plink2"
  ldsc_dir: ""
  ldsc_python: "python"
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE, error = FALSE, warning = TRUE,
                     message = FALSE, cache = FALSE, fig.width = 7, fig.height = 4.5)
library(data.table)
project_dir <- normalizePath(getwd())
source(file.path(project_dir, "R", "tutorial_helpers.R"))
source(file.path(project_dir, "R", "prs_helpers.R"))
context <- initialize_tutorial(params, project_dir)
list2env(context, envir = environment())
plink_memory_mb <- as.integer(params$plink_memory_mb)
stopifnot(!is.na(plink_memory_mb), plink_memory_mb >= 640L)

run_ldsc <- function(script, args, label, expected = character()) {
  if (params$tool_mode == "biowulf_modules") {
    arguments <- c(script, args)
  } else {
    arguments <- c(file.path(ldsc_dir, script), args)
  }
  run_command(ldsc_launcher, arguments, file.path(run_dir, label), expected)
}

section_note <- function(section) {
  if (!isTRUE(params$execute)) {
    cat("**Code preview.** This section has not been executed.\n\n")
  } else if (!can_run[[section]]) {
    cat("**Not run.** ")
    if (!context$status$requested[context$status$section == section]) {
      cat("This section was disabled in the document parameters.\n\n")
    } else {
      cat("A required input, software environment, or preceding analysis is unavailable. ",
          "See the readiness table and files in the results directory.\n\n", sep = "")
      missing <- inventory$path[inventory$section == section & !inventory$readable]
      if (length(missing)) cat(paste0("- `", head(missing, 6), "`", collapse = "\n"), "\n\n")
      if (length(missing) > 6) cat("Additional missing paths are listed in input_inventory.csv.\n\n")
    }
  } else {
    cat("This section uses the supplied data and records newly computed results.\n\n")
  }
}
```

## Purpose and preparation

This practical examines SNP heritability, shared genetic effects, functional
annotation enrichment, and polygenic scores. The analyses use GCTA/GREML,
LD Score regression, genetic correlation, stratified LD Score regression, and
clumping and thresholding. R runs the command-line programs and summarizes
their results.

Open `05_Heritability_PRS.Rmd` in **Biowulf OnDemand RStudio**. Keep the `R/`
and `scripts/` folders beside this file. Use a compute allocation with at least
2 CPUs; 16 GB RAM is a suggested starting request. Runtime and memory use have
not yet been measured on the intended allocation. The R packages `data.table`,
`knitr`, and `rmarkdown` must be installed before class. The supplied scripts
load Biowulf's GCTA, LDSC, PLINK 1.9, and PLINK 2 modules separately for each
command.

Use **Knit** to run the document and generate an HTML report. To run individual
chunks, first select **Session > Set Working Directory > To Source File Location**
in RStudio, then run the setup chunk and the subsequent chunks in order. You
can also render the document from its containing directory:

```r
rmarkdown::render("05_Heritability_PRS.Rmd")
```

The default is to execute every analysis. For a 45–60 minute class, an
instructor may assign genetic correlation and stratified LDSC as follow-up
exercises or prepare results in advance.

The document parameters control the input path, thread count, sections to run,
and output location. Setting `execute = FALSE` makes a **code-only preview**:

```r
rmarkdown::render("05_Heritability_PRS.Rmd",
                  params = list(execute = FALSE),
                  output_file = "05_Heritability_PRS.preview.html")
```

```{r preview-label, echo=FALSE, results='asis'}
if (!isTRUE(params$execute)) {
  cat("**This HTML is a code-only preview. No GCTA, LDSC, or PLINK analyses were run.**\n\n")
}
```

### Input files and results directory

The GCTA phenotype file is `phenotype.phen` in the `result` directory beside
the shared `data` directory. Leaving `gcta_phenotype` empty selects this file
automatically. The paths below identify the input data, phenotype file, and
output directory:

```{r show-paths}
data_dir
phenotype_path
run_dir
```

Each execution creates a new results subdirectory. Inputs remain read-only.
Command arguments, standard output, standard error, exit status, parameters,
input sizes, and the R session information are retained there. A failed command
stops rendering rather than displaying an earlier result as a new result.

```{r readiness, echo=FALSE}
if (isTRUE(params$execute)) {
  knitr::kable(context$status, caption = "Readiness of the requested analyses")
} else {
  cat("Preview mode does not test the Biowulf software environment.\n")
}
```

If a requested section is unavailable, its code remains visible and it is
marked **Not run**. A missing software module is distinct from a missing input;
consult `tool-*.stderr.txt` and `input_inventory.csv` in the results directory.

## SNP heritability with GCTA/GREML

The chromosome 22 data were generated from European 1000 Genomes genotypes,
with generating heritability 0.2 and 5% causal variants.
The supplied files contain 498 individuals and 34,386 variants. The generating
value is a simulation setting; the fitted estimate has sampling uncertainty.

```{r gcta-status, echo=FALSE, results='asis'}
section_note("gcta")
```

### Check phenotype identifiers

The phenotype file has three columns without a header: family ID, individual
ID, and quantitative phenotype. Both IDs must match the genotype file. The
phenotype column in `chr22.fam` contains only the missing-value code `-9` and
must not replace this phenotype file.

```{r gcta-check-phenotype, eval=can_run[['gcta']]}
phenotype_summary <- validate_gcta_phenotype(
  file.path(data_dir, "chr22.fam"), phenotype_path
)
knitr::kable(phenotype_summary)
```

### Estimate the genetic relationship matrix

```{r gcta-grm, eval=can_run[['gcta']]}
grm_prefix <- file.path(run_dir, "chr22")
run_command(
  tool_paths[["gcta"]],
  c("--bfile", file.path(data_dir, "chr22"),
    "--make-grm", "--thread-num", as.character(threads), "--out", grm_prefix),
  log_prefix = file.path(run_dir, "gcta_make_grm"),
  expected = paste0(grm_prefix, c(".grm.bin", ".grm.N.bin", ".grm.id"))
)
```

### Estimate the variance components

Apply a relatedness cutoff of 0.05. Inspect the GCTA log for the number of
individuals retained after this cutoff.

```{r gcta-reml, eval=can_run[['gcta']]}
gcta_prefix <- file.path(run_dir, "gcta_herit")
run_command(
  tool_paths[["gcta"]],
  c("--reml", "--grm", grm_prefix, "--pheno", phenotype_path,
    "--grm-cutoff", "0.05", "--thread-num", as.character(threads),
    "--out", gcta_prefix),
  log_prefix = file.path(run_dir, "gcta_reml"),
  expected = paste0(gcta_prefix, c(".hsq", ".log"))
)
cat(paste(readLines(paste0(gcta_prefix, ".hsq")), collapse = "\n"))
```

In the `.hsq` output, `V(G)` is the fitted genetic variance, `V(e)` the residual
variance, and `V(G)/Vp` their heritability ratio. Read the ratio together with
its standard error. Compare it with the generating value of 0.2 and consider
the small sample size and restriction to chromosome 22. See the
[GCTA documentation](https://yanglab.westlake.edu.cn/software/gcta/).


## LD Score regression for breast cancer

This analysis uses GWAS summary statistics for overall breast cancer
(`overall_bc`), luminal A breast cancer (`lua_bc`), and triple-negative breast
cancer (`tn_bc`). The study background is described by
[Zhang et al. (2020)](https://doi.org/10.1038/s41588-020-0609-2).
These analyses use the supplied summary statistics and European LD reference
files directly.

The primary software reference is the original authors'
[LDSC repository](https://github.com/bulik/ldsc), with instructions for
[heritability and genetic correlation](https://github.com/bulik/ldsc/wiki/Heritability-and-Genetic-Correlation)
and [partitioned heritability](https://github.com/bulik/ldsc/wiki/Partitioned-Heritability).
The supplied launcher loads the Biowulf LDSC module. A custom installation
can instead be selected with the `tool_mode`, `ldsc_dir`, and `ldsc_python`
parameters. The authors' repository also links to the
[CBIIT Python 3 implementation](https://github.com/CBIIT/ldsc).

```{r ldsc-status, echo=FALSE, results='asis'}
section_note("ldsc")
```

### Inspect the supplied summary statistics

| Column | Meaning in the supplied data |
|:--|:--|
| `snpid` | rs identifier when available; otherwise chromosome and position |
| `CHR`, `bp` | Chromosome and GRCh37 position |
| `A1`, `A2` | Effect and non-effect alleles |
| `Z`, `P` | Signed association statistic and P value |
| `info`, `MAF` | Imputation quality and minor allele frequency |
| `N` | Supplied effective sample size, calculated as \(1 / [\operatorname{var}(\hat\beta)\,2f(1-f)]\) |

The supplied `N` definition is used to estimate genetic variance on the logit
scale. The commands below supply neither population prevalence nor sample
prevalence for a liability-scale transformation. Interpret the fitted variance
in the context of this `N` definition; do not report it automatically as
liability-scale heritability or a proportion of an individual's disease risk.

Read a few rows for inspection without loading the complete overall breast
cancer file into R. LDSC will read the full file in the next step.

```{r ldsc-inspect-input, eval=can_run[['ldsc']]}
bcac_overall <- data.table::fread(
  file.path(data_dir, "overall_bc"), nrows = 6L
)
knitr::kable(bcac_overall)
```

### Prepare overall breast cancer summary statistics

Require imputation quality of at least 0.3 and minor allele frequency of at
least 0.01. Alleles are matched against the supplied
HapMap3 SNP list. Examine the munging log to see how many variants remain after
each filter.

```{r ldsc-munge-overall, eval=can_run[['ldsc']]}
munge_overall_prefix <- file.path(run_dir, "ldsc_herit_overall")
munge_result <- paste0(munge_overall_prefix, ".sumstats.gz")

run_ldsc(
  "munge_sumstats.py",
  c(
    "--sumstats", file.path(data_dir, "overall_bc"),
    "--out", munge_overall_prefix,
    "--merge-alleles", file.path(data_dir, "eur_w_ld_chr", "w_hm3.snplist"),
    "--chunksize", "500000",
    "--signed-sumstats", "Z,0",
    "--info-min", "0.3",
    "--maf-min", "0.01"
  ),
  label = "munge_overall",
  expected = c(munge_result, paste0(munge_overall_prefix, ".log"))
)
read_ldsc_log(munge_overall_prefix)
```

### Estimate the overall breast cancer genetic variance

The same chromosome-specific European LD scores are used for the reference
and regression weights. The trailing slash is required because LDSC appends
chromosome numbers to this prefix.

```{r ldsc-overall-h2, eval=can_run[['ldsc']]}
eur_ld_prefix <- paste0(file.path(data_dir, "eur_w_ld_chr"), "/")
h2_overall_prefix <- file.path(run_dir, "h2_overall")

run_ldsc(
  "ldsc.py",
  c(
    "--h2", munge_result,
    "--ref-ld-chr", eur_ld_prefix,
    "--w-ld-chr", eur_ld_prefix,
    "--out", h2_overall_prefix
  ),
  label = "estimate_overall_h2",
  expected = paste0(h2_overall_prefix, ".log")
)
read_ldsc_log(h2_overall_prefix)
```

Read the estimate together with its standard error, the mean chi-square
statistic, and the intercept. Record the scale implied by the supplied sample
size definition.

### Estimate the genetic correlation between subtypes

Prepare luminal A and triple-negative breast cancer summary statistics with
the same filters used for overall breast cancer.

```{r ldsc-inspect-subtypes, eval=can_run[['ldsc']]}
lua <- data.table::fread(file.path(data_dir, "lua_bc"), nrows = 6L)
tn <- data.table::fread(file.path(data_dir, "tn_bc"), nrows = 6L)
knitr::kable(lua, caption = "Luminal A summary statistics")
knitr::kable(tn, caption = "Triple-negative summary statistics")
```

```{r ldsc-munge-subtypes, eval=can_run[['ldsc']]}
subtype_inputs <- c(lua = "lua_bc", tn = "tn_bc")
for (subtype in names(subtype_inputs)) {
  subtype_prefix <- file.path(run_dir, paste0("ldsc_herit_", subtype))
  run_ldsc(
    "munge_sumstats.py",
    c(
      "--sumstats", file.path(data_dir, subtype_inputs[[subtype]]),
      "--out", subtype_prefix,
      "--merge-alleles", file.path(data_dir, "eur_w_ld_chr", "w_hm3.snplist"),
      "--chunksize", "500000",
      "--signed-sumstats", "Z,0",
      "--info-min", "0.3",
      "--maf-min", "0.01"
    ),
    label = paste0("munge_", subtype),
    expected = paste0(subtype_prefix, c(".sumstats.gz", ".log"))
  )
  read_ldsc_log(subtype_prefix)
}

munge_result_lua <- file.path(run_dir, "ldsc_herit_lua.sumstats.gz")
munge_result_tn <- file.path(run_dir, "ldsc_herit_tn.sumstats.gz")
```

```{r ldsc-genetic-correlation, eval=can_run[['ldsc']]}
rg_prefix <- file.path(run_dir, "rg_lua_tn")
run_ldsc(
  "ldsc.py",
  c(
    "--rg", paste(munge_result_lua, munge_result_tn, sep = ","),
    "--ref-ld-chr", eur_ld_prefix,
    "--w-ld-chr", eur_ld_prefix,
    "--out", rg_prefix
  ),
  label = "estimate_genetic_correlation",
  expected = paste0(rg_prefix, ".log")
)
read_ldsc_log(rg_prefix)
```

Report the genetic correlation with its standard error. Discuss whether the
estimate suggests shared common-variant effects between the two subtypes.
Genetic correlation alone does not establish a causal relationship.

## Stratified LD Score regression

Stratified LD Score regression estimates how genetic variance is distributed
across functional annotations. This analysis uses the baseline-LD model with
overlapping annotations, the supplied allele-frequency files and regression
weights, and the overall breast cancer summary statistics prepared above.

```{r sldsc-status, echo=FALSE, results='asis'}
section_note("sldsc")
```

```{r sldsc-estimate, eval=can_run[['sldsc']]}
sldsc_prefix <- file.path(run_dir, "h2_sldsc")
run_ldsc(
  "ldsc.py",
  c(
    "--h2", file.path(run_dir, "ldsc_herit_overall.sumstats.gz"),
    "--ref-ld-chr", file.path(
      data_dir, "1000G_Phase3_baselineLD_ldscores", "baselineLD."
    ),
    "--w-ld-chr", file.path(
      data_dir, "1000G_Phase3_weights_hm3_no_MHC", "weights.hm3_noMHC."
    ),
    "--overlap-annot",
    "--frqfile-chr", file.path(
      data_dir, "1000G_Phase3_frq", "1000G.EUR.QC."
    ),
    "--out", sldsc_prefix
  ),
  label = "estimate_stratified_h2",
  expected = paste0(sldsc_prefix, c(".results", ".log"))
)
read_ldsc_log(sldsc_prefix)
```

### Inspect annotation enrichment and its uncertainty

For a binary annotation, enrichment compares its estimated proportion of
genetic variance with its proportion of SNPs. The largest point estimate may
also have substantial uncertainty. Inspect its standard error and P value
before drawing a conclusion; the maximum is selected after comparing several
annotations.

```{r sldsc-results, eval=can_run[['sldsc']]}
enrichment_result <- data.table::fread(paste0(sldsc_prefix, ".results"))
required_enrichment_columns <- c(
  "Category", "Prop._SNPs", "Prop._h2", "Prop._h2_std_error",
  "Enrichment", "Enrichment_std_error", "Enrichment_p"
)
stopifnot(all(required_enrichment_columns %in% names(enrichment_result)))
knitr::kable(head(enrichment_result))

finite_enrichment <- which(is.finite(enrichment_result$Enrichment))
if (length(finite_enrichment) == 0L) {
  stop("The stratified LDSC results contain no finite enrichment estimates.")
}
max_enrichment_row <- finite_enrichment[
  which.max(enrichment_result$Enrichment[finite_enrichment])
]
knitr::kable(
  enrichment_result[max_enrichment_row],
  caption = "Annotation with the largest observed enrichment estimate"
)
```

Examine the Andersson enhancer annotation in more detail. Select it by name
to identify the same annotation regardless of the results table's row order.

```{r sldsc-andersson-enhancers, eval=can_run[['sldsc']]}
andersson_result <- enrichment_result[
  enrichment_result$Category == "Enhancer_AnderssonL2_0"
]
if (nrow(andersson_result) != 1L) {
  stop(
    "Expected one Enhancer_AnderssonL2_0 result. ",
    "Check the baseline annotation version and Category column."
  )
}
knitr::kable(andersson_result, caption = "Andersson enhancer annotation")
```

Compare the enhancer estimate with its uncertainty. An enrichment estimate
above one is not, by itself, evidence of statistically significant enrichment.
Interpret the overlapping annotations jointly rather than adding their
estimated proportions.

## Polygenic scores by clumping and thresholding

This exercise uses simulated chromosome 22 data, generated using
1000 Genomes European reference data. `EUR_sum_data` provides association
statistics, `1kg_eur_22/chr_22` is the LD reference, and
`prs_genotype/chr22_test` contains the target genotypes. The target outcome is
`y_out`. These inputs are separate from the breast cancer summary statistics
used above.

We use PLINK 1.9 for clumping, PLINK 2 for scoring, and R for outcome regression.
We compare nine P-value thresholds and match sample IDs explicitly before
analysis.

```{r prs-status, echo=FALSE, results='asis'}
section_note("prs")
```

### Check the association statistics and target sample IDs

Use the first 10,000 rows of `y_out` as tuning samples and the next 10,000 rows
as validation samples. Record this partition in `prs_sample_partition.tsv`.
Scores are matched to these samples by ID, so reordering a score file does not
change the partition.

```{r prs-inputs, eval=can_run[['prs']]}
sum_data_file <- file.path(data_dir, "EUR_sum_data")
reference_prefix <- file.path(data_dir, "1kg_eur_22", "chr_22")
target_prefix <- file.path(data_dir, "prs_genotype", "chr22_test")

eur_sum <- fread(sum_data_file, colClasses = c(SNP = "character", A1 = "character"))
prs_require_columns(eur_sum, c("SNP", "A1", "BETA", "P"), "EUR_sum_data")
prs_check_ids(eur_sum$SNP, "EUR_sum_data SNP")
eur_sum[, BETA := prs_as_numeric(BETA, "EUR_sum_data BETA")]
eur_sum[, P := prs_as_numeric(P, "EUR_sum_data P")]
if (any(eur_sum$P < 0 | eur_sum$P > 1)) {
  stop("EUR_sum_data contains P-values outside [0, 1].")
}
if (anyNA(eur_sum$A1) || any(!nzchar(eur_sum$A1))) {
  stop("EUR_sum_data contains a missing effect allele.")
}
if ("TEST" %in% names(eur_sum) && any(is.na(eur_sum$TEST) | eur_sum$TEST != "ADD")) {
  stop("This exercise expects one additive association result per variant.")
}

target_fam <- prs_read_fam(paste0(target_prefix, ".fam"))
target_bim <- prs_read_bim(paste0(target_prefix, ".bim"))
outcome <- prs_read_outcomes(file.path(data_dir, "y_out"), target_fam)
fwrite(outcome[, .(ID, original_row, subset)],
       file.path(run_dir, "prs_sample_partition.tsv"), sep = "\t")
knitr::kable(outcome[, .(samples = .N), by = subset])
head(eur_sum)
```

### Select index variants using the LD reference

The clumping radius is 500 kb, the LD threshold is $r^2 = 0.1$, and candidate
index variants may have P-values up to 1. See the
[PLINK clumping documentation](https://www.cog-genomics.org/plink/1.9/postproc#clump)
for the interpretation of these parameters.

```{r prs-clump, eval=can_run[['prs']]}
clump_prefix <- file.path(run_dir, "LD_clump")
run_command(
  tool_paths[["plink"]],
  c("--bfile", reference_prefix,
    "--clump", sum_data_file,
    "--clump-p1", "1", "--clump-r2", "0.1", "--clump-kb", "500",
    "--threads", as.character(threads),
    "--memory", as.character(plink_memory_mb), "--out", clump_prefix),
  log_prefix = paste0(clump_prefix, "_command"),
  expected = paste0(clump_prefix, ".clumped")
)

ld_clump <- fread(paste0(clump_prefix, ".clumped"),
                  colClasses = c(SNP = "character"))
prs_require_columns(ld_clump, "SNP", "PLINK clump output")
prs_check_ids(ld_clump$SNP, "Clumped SNP")
if (!all(ld_clump$SNP %in% eur_sum$SNP)) {
  stop("A clumped index variant has no matching association statistic.")
}
prs_prep <- eur_sum[match(ld_clump$SNP, eur_sum$SNP), .(SNP, A1, BETA, P)]
setorder(prs_prep, P, SNP)
head(prs_prep)
```

### Calculate a score at each P-value threshold

Each score sums effect-allele dosage multiplied by its estimated coefficient.
With the `no-mean-imputation` option, missing genotypes do not receive a mean
dosage contribution. The requested output column, `SCORE1_SUM`, contains the
score sum. See the [PLINK 2 scoring documentation](https://www.cog-genomics.org/plink/2.0/score).

The table records both selected index variants and variants present in the
target genotypes. PLINK skips absent variants; an allele mismatch is an error
requiring investigation. A threshold with no available variants or a constant
tuning score cannot be fitted and is excluded from threshold selection.

```{r prs-score, eval=can_run[['prs']]}
p_thresholds <- c(5e-8, 5e-7, 5e-6, 5e-5, 5e-4, 5e-3, 5e-2, 0.5, 1)
tuning_results <- data.table(
  threshold_index = seq_along(p_thresholds),
  p_threshold = p_thresholds,
  selected_variants = 0L,
  target_variants = 0L,
  tuning_r_squared = NA_real_,
  status = "not calculated"
)

for (k in seq_along(p_thresholds)) {
  coefficients <- prs_prep[P <= p_thresholds[k], .(SNP, A1, BETA)]
  tuning_results[k, selected_variants := nrow(coefficients)]
  target_index <- match(coefficients$SNP, target_bim$SNP)
  present <- !is.na(target_index)
  tuning_results[k, target_variants := sum(present)]

  if (!any(present)) {
    tuning_results[k, status := "no selected variants in target genotypes"]
    next
  }
  reference_allele <- coefficients$A1[present]
  target_alleles <- target_bim[target_index[present]]
  if (any(reference_allele != target_alleles$A1 &
          reference_allele != target_alleles$A2)) {
    stop("An effect allele does not match either target allele at threshold ",
         p_thresholds[k], ". Check allele coding before scoring.")
  }

  coefficient_file <- file.path(run_dir, paste0("prs_coeff_", k, ".tsv"))
  score_prefix <- file.path(run_dir, paste0("prs_", k))
  fwrite(coefficients, coefficient_file, sep = "\t", quote = FALSE)
  run_command(
    tool_paths[["plink2"]],
    c("--bfile", target_prefix,
      "--score", coefficient_file, "1", "2", "3", "header",
      "no-mean-imputation", "cols=+scoresums,-scoreavgs",
      "--threads", as.character(threads),
    "--memory", as.character(plink_memory_mb), "--out", score_prefix),
    log_prefix = paste0(score_prefix, "_command"),
    expected = paste0(score_prefix, ".sscore")
  )

  aligned_scores <- prs_align_scores(paste0(score_prefix, ".sscore"),
                                     outcome, target_fam)
  tuning <- aligned_scores[subset == "tuning"]
  if (var(tuning$score) == 0) {
    tuning_results[k, status := "constant tuning score"]
    next
  }
  tuning_model <- lm(y ~ score, data = tuning)
  tuning_results[k, `:=`(
    tuning_r_squared = summary(tuning_model)$r.squared,
    status = "fitted"
  )]
}

fwrite(tuning_results, file.path(run_dir, "prs_tuning_results.tsv"), sep = "\t")
knitr::kable(tuning_results, digits = 5)
```

### Select the threshold using the tuning samples

Select the threshold with the largest tuning $R^2$. If thresholds tie, the
first one in the prespecified order is selected. Validation outcomes do not
enter threshold selection.

```{r prs-select-threshold, eval=can_run[['prs']]}
eligible <- which(is.finite(tuning_results$tuning_r_squared))
if (!length(eligible)) stop("No P-value threshold produced an estimable tuning model.")
best_index <- eligible[which.max(tuning_results$tuning_r_squared[eligible])]
selected_threshold <- tuning_results[best_index]
knitr::kable(selected_threshold, digits = 5)

plot(tuning_results$p_threshold, tuning_results$tuning_r_squared,
     log = "x", type = "b", pch = 16,
     xlab = "P-value threshold", ylab = expression(Tuning~R^2))
points(selected_threshold$p_threshold, selected_threshold$tuning_r_squared,
       pch = 21, bg = "white", cex = 1.6)
```

### Evaluate the selected score in the validation samples

Evaluate the association between the selected score and the outcome using
$R^2$ from `lm(y ~ score)` fitted within the validation samples. The threshold
and variant weights are fixed before this step, but the regression intercept
and slope are estimated in the validation samples. Thus this is not a measure
of calibration or prediction error for an outcome model fitted entirely in
the tuning samples.

```{r prs-validation, eval=can_run[['prs']]}
selected_scores <- prs_align_scores(
  file.path(run_dir, paste0("prs_", best_index, ".sscore")), outcome, target_fam
)
validation <- selected_scores[subset == "validation"]
if (var(validation$score) == 0) {
  stop("The selected score is constant in the validation samples; R-squared is undefined.")
}
validation_model <- lm(y ~ score, data = validation)
validation_results <- data.table(
  selected_p_threshold = p_thresholds[best_index],
  tuning_n = sum(outcome$subset == "tuning"),
  validation_n = nrow(validation),
  tuning_r_squared = selected_threshold$tuning_r_squared,
  validation_r_squared = summary(validation_model)$r.squared
)
fwrite(validation_results, file.path(run_dir, "prs_validation_results.tsv"), sep = "\t")
saveRDS(validation_model, file.path(run_dir, "prs_validation_model.rds"))
knitr::kable(validation_results, digits = 5)
```

**Interpretation.** Compare the tuning and validation estimates. A larger tuning
$R^2$ may reflect selection among thresholds. These results describe this
simulated chromosome 22 exercise; they do not establish genome-wide predictive
performance or transferability to another population.


## Analysis record

```{r record-analysis, echo=FALSE}
final_status <- context$status
final_status$status[can_run] <- "completed"
if (!isTRUE(params$execute)) final_status$status <- "preview only; not executed"
write.csv(final_status, file.path(run_dir, "section_status.csv"), row.names = FALSE)
writeLines(capture.output(sessionInfo()), file.path(run_dir, "sessionInfo.txt"))
knitr::kable(final_status[, c("section", "status")])
cat("Results directory:", run_dir, "\n")
```

```{r session-information, echo=FALSE}
sessionInfo()
```
