---
title: "Heritability"
subtitle: "Statistical genetics practical in RStudio on Biowulf"
output:
  html_document:
    toc: true
    toc_float: false
    number_sections: true
    code_folding: show
    self_contained: true
    theme: readable
params:
  execute: true
  data_dir: "/data/DCEG_shared/statgen_workshop_2026/data/statgen_workshop_04Heritability_PRS/data"
  gcta_phenotype: ""
  output_dir: "results"
  threads: 1
  run_gcta: true
  run_ldsc: true
  run_sldsc: true
  tool_mode: "biowulf_modules"
  gcta_bin: "gcta64"
  ldsc_dir: ""
  ldsc_python: "python"
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE, error = FALSE, warning = TRUE,
                     message = FALSE, cache = FALSE, fig.width = 7, fig.height = 4.5)
library(data.table)
project_dir <- normalizePath(getwd())
source(file.path(project_dir, "R", "tutorial_helpers.R"))
context <- initialize_tutorial(params, project_dir)
list2env(context, envir = environment())

run_ldsc <- function(script, args, label, expected = character()) {
  if (params$tool_mode == "biowulf_modules") {
    arguments <- c(script, args)
  } else {
    arguments <- c(file.path(ldsc_dir, script), args)
  }
  run_command(ldsc_launcher, arguments, file.path(run_dir, label), expected)
}

section_note <- function(section) {
  if (!isTRUE(params$execute)) {
    cat("**Code preview.** This section has not been executed.\n\n")
  } else if (!can_run[[section]]) {
    cat("**Not run.** ")
    if (!context$status$requested[context$status$section == section]) {
      cat("This section was disabled in the document parameters.\n\n")
    } else {
      cat("A required input, software environment, or preceding analysis is unavailable. ",
          "See the readiness table and files in the results directory.\n\n", sep = "")
      missing <- inventory$path[inventory$section == section & !inventory$readable]
      if (length(missing)) cat(paste0("- `", head(missing, 6), "`", collapse = "\n"), "\n\n")
      if (length(missing) > 6) cat("Additional missing paths are listed in input_inventory.csv.\n\n")
    }
  } else {
    cat("This section uses the supplied data and records newly computed results.\n\n")
  }
}
```

## Purpose and preparation

This practical examines SNP heritability, shared genetic effects, and functional
annotation enrichment. The analyses use GCTA/GREML, LD Score regression,
genetic correlation, and stratified LD Score regression. R runs the command-line
programs and summarizes their results.

Start **Biowulf OnDemand RStudio** with a compute allocation of at least
2 CPUs; 16 GB RAM is a suggested starting request. Runtime and memory use have
not yet been measured on the intended allocation. The R packages `data.table`,
`knitr`, and `rmarkdown` must be installed before class. The supplied scripts
load Biowulf's GCTA and LDSC modules separately for each command.

### Clone and open your own copy

In RStudio, open the **Terminal** tab and run the commands below. These are
shell commands, so enter them in the Terminal rather than the R Console.
Clone the repository into a folder you own; this example uses
`/data/$USER/Stats_Gen/workshop5`, where `$USER` expands to your Biowulf username.
You can substitute another writable folder of your own.

```bash
mkdir -p "/data/$USER/Stats_Gen/workshop5"
cd "/data/$USER/Stats_Gen/workshop5"
git clone https://github.com/DCEG-workshops/statgen_workshop_tutorial.git
cd statgen_workshop_tutorial/src/05_Heritability
pwd
```

In RStudio, select **File > Open File...** and open `05_Heritability.Rmd`
from the folder printed by `pwd`. With the example above, the full path is:

```text
/data/<username>/Stats_Gen/workshop5/statgen_workshop_tutorial/src/05_Heritability/05_Heritability.Rmd
```

Replace `<username>` with your Biowulf username. Keep the `R/` and `scripts/`
folders beside the Rmd file; cloning the repository includes both. If you
already cloned the repository into your own folder, open the Rmd from that
copy instead of cloning again.

### Run the tutorial

Use **Knit** to run the document and generate an HTML report. To run individual
chunks, first select **Session > Set Working Directory > To Source File Location**
in RStudio, then run the setup chunk and the subsequent chunks in order.
The Terminal's `cd` command does not change the R session's working directory.
You can also render the document from its containing directory:

```r
rmarkdown::render("05_Heritability.Rmd")
```

The default is to execute every analysis. For a 45–60 minute class, an
instructor may assign genetic correlation and stratified LDSC as follow-up
exercises or prepare results in advance.

The document parameters control the input path, thread count, sections to run,
and output location. Setting `execute = FALSE` makes a **code-only preview**:

```r
rmarkdown::render("05_Heritability.Rmd",
                  params = list(execute = FALSE),
                  output_file = "05_Heritability.preview.html")
```

```{r preview-label, echo=FALSE, results='asis'}
if (!isTRUE(params$execute)) {
  cat("**This HTML is a code-only preview. No GCTA or LDSC analyses were run.**\n\n")
}
```

### Input files and results directory

The GCTA phenotype file is `phenotype.phen` in the `result` directory beside
the shared `data` directory. Leaving `gcta_phenotype` empty selects this file
automatically. The paths below identify the input data, phenotype file, and
output directory:

```{r show-paths}
data_dir
phenotype_path
run_dir
```

Each execution creates a new results subdirectory. Inputs remain read-only.
Command arguments, standard output, standard error, exit status, parameters,
input sizes, and the R session information are retained there. A failed command
stops rendering rather than displaying an earlier result as a new result.

```{r readiness, echo=FALSE}
if (isTRUE(params$execute)) {
  knitr::kable(context$status, caption = "Readiness of the requested analyses")
} else {
  cat("Preview mode does not test the Biowulf software environment.\n")
}
```

If a requested section is unavailable, its code remains visible and it is
marked **Not run**. A missing software module is distinct from a missing input;
consult `tool-*.stderr.txt` and `input_inventory.csv` in the results directory.

## SNP heritability with GCTA/GREML

The chromosome 22 data were generated from European 1000 Genomes genotypes,
with generating heritability 0.2 and 5% causal variants.
The supplied files contain 498 individuals and 34,386 variants. The generating
value is a simulation setting; the fitted estimate has sampling uncertainty.

```{r gcta-status, echo=FALSE, results='asis'}
section_note("gcta")
```

### Check phenotype identifiers

The phenotype file has three columns without a header: family ID, individual
ID, and quantitative phenotype. Both IDs must match the genotype file. The
phenotype column in `chr22.fam` contains only the missing-value code `-9` and
must not replace this phenotype file.

```{r gcta-check-phenotype, eval=can_run[['gcta']]}
phenotype_summary <- validate_gcta_phenotype(
  file.path(data_dir, "chr22.fam"), phenotype_path
)
knitr::kable(phenotype_summary)
```

### Estimate the genetic relationship matrix

```{r gcta-grm, eval=can_run[['gcta']]}
grm_prefix <- file.path(run_dir, "chr22")
run_command(
  tool_paths[["gcta"]],
  c("--bfile", file.path(data_dir, "chr22"),
    "--make-grm", "--thread-num", as.character(threads), "--out", grm_prefix),
  log_prefix = file.path(run_dir, "gcta_make_grm"),
  expected = paste0(grm_prefix, c(".grm.bin", ".grm.N.bin", ".grm.id"))
)
```

### Estimate the variance components

Apply a relatedness cutoff of 0.05. Inspect the GCTA log for the number of
individuals retained after this cutoff.

```{r gcta-reml, eval=can_run[['gcta']]}
gcta_prefix <- file.path(run_dir, "gcta_herit")
run_command(
  tool_paths[["gcta"]],
  c("--reml", "--grm", grm_prefix, "--pheno", phenotype_path,
    "--grm-cutoff", "0.05", "--thread-num", as.character(threads),
    "--out", gcta_prefix),
  log_prefix = file.path(run_dir, "gcta_reml"),
  expected = paste0(gcta_prefix, c(".hsq", ".log"))
)
cat(paste(readLines(paste0(gcta_prefix, ".hsq")), collapse = "\n"))
```

In the `.hsq` output, `V(G)` is the fitted genetic variance, `V(e)` the residual
variance, and `V(G)/Vp` their heritability ratio. Read the ratio together with
its standard error. Compare it with the generating value of 0.2 and consider
the small sample size and restriction to chromosome 22. See the
[GCTA documentation](https://yanglab.westlake.edu.cn/software/gcta/).


## LD Score regression for breast cancer

This analysis uses GWAS summary statistics for overall breast cancer
(`overall_bc`), luminal A breast cancer (`lua_bc`), and triple-negative breast
cancer (`tn_bc`). The study background is described by
[Zhang et al. (2020)](https://doi.org/10.1038/s41588-020-0609-2).
These analyses use the supplied summary statistics and European LD reference
files directly.

The primary software reference is the original authors'
[LDSC repository](https://github.com/bulik/ldsc), with instructions for
[heritability and genetic correlation](https://github.com/bulik/ldsc/wiki/Heritability-and-Genetic-Correlation)
and [partitioned heritability](https://github.com/bulik/ldsc/wiki/Partitioned-Heritability).
The supplied launcher loads the Biowulf LDSC module. A custom installation
can instead be selected with the `tool_mode`, `ldsc_dir`, and `ldsc_python`
parameters. The authors' repository also links to the
[CBIIT Python 3 implementation](https://github.com/CBIIT/ldsc).

```{r ldsc-status, echo=FALSE, results='asis'}
section_note("ldsc")
```

### Inspect the supplied summary statistics

| Column | Meaning in the supplied data |
|:--|:--|
| `snpid` | rs identifier when available; otherwise chromosome and position |
| `CHR`, `bp` | Chromosome and GRCh37 position |
| `A1`, `A2` | Effect and non-effect alleles |
| `Z`, `P` | Signed association statistic and P value |
| `info`, `MAF` | Imputation quality and minor allele frequency |
| `N` | Supplied effective sample size, calculated as \(1 / [\operatorname{var}(\hat\beta)\,2f(1-f)]\) |

The supplied `N` definition is used to estimate genetic variance on the logit
scale. The commands below supply neither population prevalence nor sample
prevalence for a liability-scale transformation. Interpret the fitted variance
in the context of this `N` definition; do not report it automatically as
liability-scale heritability or a proportion of an individual's disease risk.

Read a few rows for inspection without loading the complete overall breast
cancer file into R. LDSC will read the full file in the next step.

```{r ldsc-inspect-input, eval=can_run[['ldsc']]}
bcac_overall <- data.table::fread(
  file.path(data_dir, "overall_bc"), nrows = 6L
)
knitr::kable(bcac_overall)
```

### Prepare overall breast cancer summary statistics

Require imputation quality of at least 0.3 and minor allele frequency of at
least 0.01. Alleles are matched against the supplied
HapMap3 SNP list. Examine the munging log to see how many variants remain after
each filter.

```{r ldsc-munge-overall, eval=can_run[['ldsc']]}
munge_overall_prefix <- file.path(run_dir, "ldsc_herit_overall")
munge_result <- paste0(munge_overall_prefix, ".sumstats.gz")

run_ldsc(
  "munge_sumstats.py",
  c(
    "--sumstats", file.path(data_dir, "overall_bc"),
    "--out", munge_overall_prefix,
    "--merge-alleles", file.path(data_dir, "eur_w_ld_chr", "w_hm3.snplist"),
    "--chunksize", "500000",
    "--signed-sumstats", "Z,0",
    "--info-min", "0.3",
    "--maf-min", "0.01"
  ),
  label = "munge_overall",
  expected = c(munge_result, paste0(munge_overall_prefix, ".log"))
)
read_ldsc_log(munge_overall_prefix)
```

### Estimate the overall breast cancer genetic variance

The same chromosome-specific European LD scores are used for the reference
and regression weights. The trailing slash is required because LDSC appends
chromosome numbers to this prefix.

```{r ldsc-overall-h2, eval=can_run[['ldsc']]}
eur_ld_prefix <- paste0(file.path(data_dir, "eur_w_ld_chr"), "/")
h2_overall_prefix <- file.path(run_dir, "h2_overall")

run_ldsc(
  "ldsc.py",
  c(
    "--h2", munge_result,
    "--ref-ld-chr", eur_ld_prefix,
    "--w-ld-chr", eur_ld_prefix,
    "--out", h2_overall_prefix
  ),
  label = "estimate_overall_h2",
  expected = paste0(h2_overall_prefix, ".log")
)
read_ldsc_log(h2_overall_prefix)
```

Read the estimate together with its standard error, the mean chi-square
statistic, and the intercept. Record the scale implied by the supplied sample
size definition.

### Estimate the genetic correlation between subtypes

Prepare luminal A and triple-negative breast cancer summary statistics with
the same filters used for overall breast cancer.

```{r ldsc-inspect-subtypes, eval=can_run[['ldsc']]}
lua <- data.table::fread(file.path(data_dir, "lua_bc"), nrows = 6L)
tn <- data.table::fread(file.path(data_dir, "tn_bc"), nrows = 6L)
knitr::kable(lua, caption = "Luminal A summary statistics")
knitr::kable(tn, caption = "Triple-negative summary statistics")
```

```{r ldsc-munge-subtypes, eval=can_run[['ldsc']]}
subtype_inputs <- c(lua = "lua_bc", tn = "tn_bc")
for (subtype in names(subtype_inputs)) {
  subtype_prefix <- file.path(run_dir, paste0("ldsc_herit_", subtype))
  run_ldsc(
    "munge_sumstats.py",
    c(
      "--sumstats", file.path(data_dir, subtype_inputs[[subtype]]),
      "--out", subtype_prefix,
      "--merge-alleles", file.path(data_dir, "eur_w_ld_chr", "w_hm3.snplist"),
      "--chunksize", "500000",
      "--signed-sumstats", "Z,0",
      "--info-min", "0.3",
      "--maf-min", "0.01"
    ),
    label = paste0("munge_", subtype),
    expected = paste0(subtype_prefix, c(".sumstats.gz", ".log"))
  )
  read_ldsc_log(subtype_prefix)
}

munge_result_lua <- file.path(run_dir, "ldsc_herit_lua.sumstats.gz")
munge_result_tn <- file.path(run_dir, "ldsc_herit_tn.sumstats.gz")
```

```{r ldsc-genetic-correlation, eval=can_run[['ldsc']]}
rg_prefix <- file.path(run_dir, "rg_lua_tn")
run_ldsc(
  "ldsc.py",
  c(
    "--rg", paste(munge_result_lua, munge_result_tn, sep = ","),
    "--ref-ld-chr", eur_ld_prefix,
    "--w-ld-chr", eur_ld_prefix,
    "--out", rg_prefix
  ),
  label = "estimate_genetic_correlation",
  expected = paste0(rg_prefix, ".log")
)
read_ldsc_log(rg_prefix)
```

Report the genetic correlation with its standard error. Discuss whether the
estimate suggests shared common-variant effects between the two subtypes.
Genetic correlation alone does not establish a causal relationship.

## Stratified LD Score regression

Stratified LD Score regression estimates how genetic variance is distributed
across functional annotations. This analysis uses the baseline-LD model with
overlapping annotations, the supplied allele-frequency files and regression
weights, and the overall breast cancer summary statistics prepared above.

```{r sldsc-status, echo=FALSE, results='asis'}
section_note("sldsc")
```

```{r sldsc-estimate, eval=can_run[['sldsc']]}
sldsc_prefix <- file.path(run_dir, "h2_sldsc")
run_ldsc(
  "ldsc.py",
  c(
    "--h2", file.path(run_dir, "ldsc_herit_overall.sumstats.gz"),
    "--ref-ld-chr", file.path(
      data_dir, "1000G_Phase3_baselineLD_ldscores", "baselineLD."
    ),
    "--w-ld-chr", file.path(
      data_dir, "1000G_Phase3_weights_hm3_no_MHC", "weights.hm3_noMHC."
    ),
    "--overlap-annot",
    "--frqfile-chr", file.path(
      data_dir, "1000G_Phase3_frq", "1000G.EUR.QC."
    ),
    "--out", sldsc_prefix
  ),
  label = "estimate_stratified_h2",
  expected = paste0(sldsc_prefix, c(".results", ".log"))
)
read_ldsc_log(sldsc_prefix)
```

### Inspect annotation enrichment and its uncertainty

For a binary annotation, enrichment compares its estimated proportion of
genetic variance with its proportion of SNPs. The largest point estimate may
also have substantial uncertainty. Inspect its standard error and P value
before drawing a conclusion; the maximum is selected after comparing several
annotations.

```{r sldsc-results, eval=can_run[['sldsc']]}
enrichment_result <- data.table::fread(paste0(sldsc_prefix, ".results"))
required_enrichment_columns <- c(
  "Category", "Prop._SNPs", "Prop._h2", "Prop._h2_std_error",
  "Enrichment", "Enrichment_std_error", "Enrichment_p"
)
stopifnot(all(required_enrichment_columns %in% names(enrichment_result)))
knitr::kable(head(enrichment_result))

finite_enrichment <- which(is.finite(enrichment_result$Enrichment))
if (length(finite_enrichment) == 0L) {
  stop("The stratified LDSC results contain no finite enrichment estimates.")
}
max_enrichment_row <- finite_enrichment[
  which.max(enrichment_result$Enrichment[finite_enrichment])
]
knitr::kable(
  enrichment_result[max_enrichment_row],
  caption = "Annotation with the largest observed enrichment estimate"
)
```

Examine the Andersson enhancer annotation in more detail. Select it by name
to identify the same annotation regardless of the results table's row order.

```{r sldsc-andersson-enhancers, eval=can_run[['sldsc']]}
andersson_result <- enrichment_result[
  enrichment_result$Category == "Enhancer_AnderssonL2_0"
]
if (nrow(andersson_result) != 1L) {
  stop(
    "Expected one Enhancer_AnderssonL2_0 result. ",
    "Check the baseline annotation version and Category column."
  )
}
knitr::kable(andersson_result, caption = "Andersson enhancer annotation")
```

Compare the enhancer estimate with its uncertainty. An enrichment estimate
above one is not, by itself, evidence of statistically significant enrichment.
Interpret the overlapping annotations jointly rather than adding their
estimated proportions.

## Analysis record

```{r record-analysis, echo=FALSE}
final_status <- context$status
final_status$status[can_run] <- "completed"
if (!isTRUE(params$execute)) final_status$status <- "preview only; not executed"
write.csv(final_status, file.path(run_dir, "section_status.csv"), row.names = FALSE)
writeLines(capture.output(sessionInfo()), file.path(run_dir, "sessionInfo.txt"))
knitr::kable(final_status[, c("section", "status")])
cat("Results directory:", run_dir, "\n")
```

```{r session-information, echo=FALSE}
sessionInfo()
```
