---
title: "Session 1: Introduction to GWAS Quality Control"
output: html_document
---

# Objectives

- Introduce the RStudio Server / Biowulf HPC environment for genetic data analysis.
- Introduce the importance of Quality Control (QC) in the GWAS analysis pipeline.
- Provide hands-on practice performing standard QC steps on real GWAS data using PLINK.

# Prerequisites

- A basic understanding of genetics, genome-wide association studies (GWAS), and the role of QC in GWAS.
- Familiarity with the command line, since we will use PLINK (a command-line tool) for QC.
- A Biowulf HPC account.
- RStudio launched from NIH OnDemand.
- **PLINK 1.9 must be loaded before starting your OnDemand RStudio session.**

# Workshop Overview

We will use PLINK to carry out several essential QC procedures on real genotype data from the PennCATH study of coronary artery disease. This dataset includes 1,401 individuals genotyped at 861,473 markers.

In this session we will:

1. Check sample and SNP missingness
2. Identify markers that deviate from Hardy-Weinberg Equilibrium
3. Check heterozygosity across samples
4. Check estimated relatedness (IBD) between samples
5. Examine population structure using PCA, including a joint analysis with 1000 Genomes (1KG) reference samples to interpret study-sample genetic similarity relative to 1KG reference ancestry labels
6. Use GrafAnc to obtain reference-based genetic ancestry assignments and compare them with the PCA patterns

**Note: These steps illustrate core GWAS QC concepts but are not an exhaustive production QC protocol.** Depending on the study and genotyping platform, additional checks may include sex-chromosome concordance, duplicate/sample identity checks, batch effects, allele/build harmonization, differential missingness, external allele-frequency comparisons, and post-imputation QC.

---

# Setup

## Set path variables

Paths are defined once in the R session below. `Sys.setenv()` makes the same values available to each subsequent `{bash}` shell.

```{r setup-paths}
username <- Sys.getenv("USER")

if (!nzchar(username)) {
  stop("USER is not set in this R session.")
}

input_dir <- "/data/DCEG_shared/statgen_workshop_2026/data/workshop1/inputs/penncath_withPheno"
reference_dir <- "/data/DCEG_shared/statgen_workshop_2026/data/workshop1/inputs/ref"
analysis_dir <- file.path(
  "/data", username, "Stats_Gen", "workshop1", "01_analysis"
)

dir.create(analysis_dir, recursive = TRUE, showWarnings = FALSE)

if (!dir.exists(analysis_dir)) {
  stop("Could not create analysis directory: ", analysis_dir)
}

Sys.setenv(
  input_dir = input_dir,
  reference_dir = reference_dir,
  analysis_dir = analysis_dir
)
```

```{bash list-input-directory}
ls ${input_dir}
```

## Confirm PLINK is available

This workshop uses **PLINK 1.9**. Confirm your module loaded correctly:

```{bash check-plink}
which plink
plink --version
```

## Load R libraries

R chunks in this document share a single persistent R session, so libraries only need to be loaded once — but we'll reload them at the top of major sections as a safety net in case you run chunks out of order.

```{r load-libraries}
library(tidyverse)
library(data.table)
library(glue)
```

---

# Data Overview

The dataset is PLINK's standard binary format: `.bed` (genotypes), `.bim` (marker info), `.fam` (sample info and phenotype).

```{bash view-fam}
head ${input_dir}/penncath.fam
```

Column 6 of the `.fam` file is the phenotype: here, coronary artery disease (CAD) case/control status (`1` = control, `2` = case, `-9` = missing).

---

# QC Step 1: Missingness

## Sample and variant missingness

We start by calculating the proportion of missing genotype calls, per variant and per sample.

```{bash run-missing}

plink --bfile ${input_dir}/penncath --missing --out ${analysis_dir}/penncath
```

This writes `penncath.lmiss` (per-variant missingness) and `penncath.imiss` (per-sample missingness).

## Variant missingness distribution

```{r plot-variant-missing}

variant_miss <- fread(file.path(analysis_dir, "penncath.lmiss"))

variant_miss %>% ggplot(aes(x = F_MISS)) +
  geom_histogram() +
  labs(title = "Variant Missing Rate") +
  theme_classic()
```

**Question:** If we filter variants with missingness > 1%, how many would be removed?

```{r count-variant-missing}
variant_miss %>%
  mutate(missing_rate_gt0.01 = if_else(F_MISS > 0.01, "Yes", "No")) %>%
  group_by(missing_rate_gt0.01) %>%
  tally()
```

## Apply the variant missingness filter

```{bash filter-geno}

plink --bfile ${input_dir}/penncath --geno 0.01 --make-bed --out ${analysis_dir}/penncath.geno0.01
```

## Re-check sample missingness on the filtered data

```{bash sample-missing}

plink --bfile ${analysis_dir}/penncath.geno0.01 --missing --out ${analysis_dir}/penncath.geno0.01
```

```{r check-sample-missing}
sample_miss <- fread(file.path(analysis_dir, "penncath.geno0.01.imiss"))

sample_miss %>%
  mutate(missing_rate_gt0.01 = if_else(F_MISS > 0.01, "Yes", "No")) %>%
  group_by(missing_rate_gt0.01) %>%
  tally()
```

## Apply the sample missingness filter

```{bash filter-mind}

plink --bfile ${analysis_dir}/penncath.geno0.01 --mind 0.01 --make-bed --out ${analysis_dir}/penncath.geno0.01.mind0.01
```

---

# QC Step 2: Hardy-Weinberg Equilibrium (HWE)

For each SNP, we test whether observed genotype frequencies deviate from what's expected under Hardy-Weinberg equilibrium. Large deviations can indicate technical artifact or genotyping error, though they may also reflect true population-genetic phenomena (e.g., selection), population stratification.

```{bash run-hwe}

plink --bfile ${analysis_dir}/penncath.geno0.01.mind0.01 --hardy --out ${analysis_dir}/penncath.geno0.01.mind0.01
```

**Important:** PLINK's `--hardy` output reports up to three rows per SNP — `ALL` (combined sample), `AFF` (cases only), and `UNAFF` (controls only) — when a case/control phenotype is present. For QC in a case/control study, we use the controls (`UNAFF`) because a genuine disease association can cause HWE departures among cases.

```{r hwe-check}
hwe_stat <- fread(file.path(analysis_dir, "penncath.geno0.01.mind0.01.hwe"))
hwe_stat <- hwe_stat %>% filter(TEST == "UNAFF")

hwe_stat %>% ggplot(aes(x = P)) +
  geom_histogram() +
  labs(title = "Hardy-Weinberg equilibrium exact test p-values") +
  theme_classic()
```

**Question:** How many markers have a HWE p-value below 1e-8?

```{r hwe-count}
hwe_stat %>%
  mutate(hwe_stat_lessThan1e_8 = if_else(P < 1e-8, TRUE, FALSE)) %>%
  group_by(hwe_stat_lessThan1e_8) %>%
  tally()
```

**Question for discussion:** What HWE p-value threshold would you use to filter markers? PLINK removes markers with HWE p-values below the selected cutoff, so a higher cutoff (e.g., 1e-3) is more aggressive and removes more markers, while a lower cutoff (e.g., 1e-6) is more permissive and retains more. In datasets with multiple ancestries, apparent HWE violations can also reflect population stratification rather than genotyping error — how might you account for this?

*(We will not apply an HWE filter within this notebook — see the exercise at the end, where you'll build your own HWE-filtered dataset.)*

---

# QC Step 3: LD Pruning

Many downstream QC analyses, including PLINK's Identity-by-Descent (IBD), heterozygosity rate, and principal component analysis (PCA), can be influenced by correlated markers in linkage disequilibrium (LD). We therefore use an approximately LD-independent marker set for these analyses so that regions of high LD do not receive disproportionate weight.

Here we use a 50-marker window, sliding 5 markers at a time, removing one marker from any pair with LD (r²) > 0.3.

```{bash ld-pruning}

plink --bfile ${analysis_dir}/penncath.geno0.01.mind0.01 --indep-pairwise 50 5 0.3 --out ${analysis_dir}/penncath.geno0.01.mind0.01
```

This produces `.prune.in` (markers to keep) and `.prune.out` (markers to remove) — we'll use `.prune.in` for the next several steps.

---

# QC Step 4: Relatedness (IBD)

We estimate pairwise relatedness (IBD) between all samples, to check for unexpected duplicates or close relatives.

```{bash ibd-check}

plink --bfile ${analysis_dir}/penncath.geno0.01.mind0.01 \
      --extract ${analysis_dir}/penncath.geno0.01.mind0.01.prune.in \
      --genome --out ${analysis_dir}/penncath.geno0.01.mind0.01.LD0.3
```

`PI_HAT` estimates the proportion of the genome shared identical-by-descent between a pair of samples: ~1.0 for duplicates/monozygotic twins, ~0.5 for parent-child or full siblings, ~0.25 for half-siblings/grandparent-grandchild/avuncular, ~0.125 for first cousins, and near 0 for unrelated individuals.

```{r ibd-summary}
ibd <- fread(file.path(analysis_dir, "penncath.geno0.01.mind0.01.LD0.3.genome"))

summary(ibd$PI_HAT)
```

We can also visualize `Z0` (probability of sharing 0 alleles IBD) vs. `Z1` (probability of sharing 1 allele IBD) for each pair. Unrelated pairs cluster near `Z0 = 1, Z1 = 0`.

```{r ibd-plot}
ibd %>% ggplot(aes(x = Z0, y = Z1)) +
  geom_point() +
  xlim(0, 1) + ylim(0, 1) +
  labs(title = "Z0 vs. Z1") +
  theme_classic()
```

**Question:** Based on the maximum PI_HAT observed, do we have any duplicate samples or close relatives that require additional handling? For an analysis that assumes unrelated individuals, that might mean removing one member of a related pair; alternatively, downstream methods that explicitly model relatedness can retain such samples.

---

# QC Step 5: Heterozygosity

Samples with unusually high or low heterozygosity can reflect technical problems (for example, contamination or genotype-calling artifacts), biological factors such as inbreeding, or differences in genetic ancestry. Heterozygosity outliers should therefore be interpreted together with other QC information rather than treated as diagnostic on their own.

```{bash het-check}

plink --bfile ${analysis_dir}/penncath.geno0.01.mind0.01 --het \
      --extract ${analysis_dir}/penncath.geno0.01.mind0.01.prune.in \
      --out ${analysis_dir}/penncath.geno0.01.mind0.01.LD0.3
```

```{r het-summary}
heterozygosity <- fread(file.path(analysis_dir, "penncath.geno0.01.mind0.01.LD0.3.het"))
heterozygosity <- heterozygosity %>% mutate(het_rate = (`N(NM)` - `O(HOM)`) / `N(NM)`)

heterozygosity %>% ggplot(aes(x = het_rate)) +
  geom_histogram() +
  labs(title = "Heterozygosity") +
  theme_classic()
```

**Question:** Which samples fall more than 3 standard deviations from the mean heterozygosity rate?

```{r het-outliers}
heterozygosity %>%
  filter(het_rate < mean(het_rate) - 3*sd(het_rate) |
         het_rate > mean(het_rate) + 3*sd(het_rate))
```

**A note on choosing a metric:** the `.het` output also provides `F`, PPLINK’s method-of-moments inbreeding coefficient estimate, alongside raw counts.  While `het_rate` measures the simple proportion of heterozygous calls, `F` contrasts observed individual homozygosity against the baseline expected from cohort allele frequencies (e.g., populations that underwent historic genetic bottlenecks aturally have lower overall baseline heterozygosity than populations with highly admixed histories). Crucially, because `F` calculations rely on a specific dataset's allele-frequency reference, uncorrected population stratification will distort its distribution. Fixed cutoffs such as `|F| > 0.1` (or, in some pipelines, `|F| > 0.2`) are sometimes used, but they are not universal; dataset-specific outlier rules such as ±3 SD, ideally evaluated within genetically comparable groups, are also common.

Try applying that threshold to this same data:

```{r het-F-threshold}
heterozygosity %>%
  filter(F > 0.1 | F < -0.1)
```

**Question:** How does this compare to the ±3 SD result above? What does this tell you about the sensitivity of QC decisions to the choice of metric and threshold?

---

# QC Step 6: Population Structure (PCA)

We use Principal Component Analysis (PCA) on common-frequency, LD-pruned markers to summarize major axes of genome-wide genetic variation. PCA does not measure ancestry directly: the resulting components can reflect genetic ancestry and demographic history, but they can also be affected by technical artifacts. For PLINK PCA, it is usually best to use markers in approximate LD; otherwise, regions of strong or long-range LD can disproportionately influence individual components. We also apply `--maf 0.05` to focus this reference-based PCA on common variants; low-frequency variants can generate sample-specific structure and are generally less useful for visualizing broad population structure in this setting.

**Terminology:** throughout this workshop, *genetic ancestry* refers to patterns of genetic similarity inferred relative to reference samples or statistical models. It is not interchangeable with self-identified race, ethnicity, nationality, or personal identity.

```{bash simple-pca}

plink --bfile ${analysis_dir}/penncath.geno0.01.mind0.01 \
      --extract ${analysis_dir}/penncath.geno0.01.mind0.01.prune.in \
      --maf 0.05 --pca --out ${analysis_dir}/penncath.geno0.01.mind0.01.LD0.3
```

```{r plot-pca}
pca <- fread(file.path(analysis_dir, "penncath.geno0.01.mind0.01.LD0.3.eigenvec"))

pca %>% rename(PC1 = V3, PC2 = V4) %>%
  ggplot(aes(x = PC1, y = PC2)) +
  geom_point() +
  labs(title = "PC1 vs. PC2") +
  theme_classic()

pca %>% rename(PC2 = V4, PC3 = V5) %>%
  ggplot(aes(x = PC2, y = PC3)) +
  geom_point() +
  labs(title = "PC2 vs. PC3") +
  theme_classic()
```

**Question:** Do you see distinct clusters in this plot? What might explain them? We'll investigate whether the pattern aligns with 1000 Genomes reference population structure below, while keeping in mind that PCA clustering can also reflect technical effects.

---

# QC Step 7: Reference-based PCA with 1000 Genomes

To help interpret the clustering seen in our study-only PCA, we merge PennCATH with 1000 Genomes (1KG) reference samples that have published population and superpopulation labels (defined initially using sampling criteria based on geography, community engagement, and strict self-reported grandparental ancestry). These labels provide a reference frame for genetic similarity on the PCs; they should not be treated as categorical ground truth for an individual's ancestry.

## Attempt the merge

`--bmerge` combines PennCATH (the base dataset) with the 1000 Genomes reference (the dataset being merged in). The result is the **union** of samples (all PennCATH + all 1000G individuals) and the **union** of variants (any marker present in either file).

```{bash merge-1kg}

plink --bfile ${analysis_dir}/penncath.geno0.01.mind0.01 \
      --bmerge ${reference_dir}/all_pops_1kgenomes_markersForPCA.bed \
      ${reference_dir}/all_pops_1kgenomes_markersForPCA.bim \
      ${reference_dir}/all_pops_1kgenomes_markersForPCA.fam \
      --make-bed --out ${analysis_dir}/penncath.geno0.01.mind0.01.merge1kgenomes || true
```

When a binary merge fails and PLINK writes a `.missnp` file, the listed variants have allele combinations that would appear to contain more than two alleles after merging. Strand-orientation mismatches are a common cause, but genuine multiallelic variants, allele-coding inconsistencies, or other data errors can produce the same symptom.

## Flip mismatched markers

`--flip` complements the allele codes (A↔T and C↔G) for the flagged variants. If most merge conflicts disappear after the flip, that supports strand orientation as the explanation for those variants. Palindromic A/T and C/G SNPs remain strand-ambiguous from allele labels alone, so a production analysis would typically resolve them with additional reference/allele-frequency information or exclude ambiguous variants when appropriate.

```{bash flip-snps}

plink --bfile ${analysis_dir}/penncath.geno0.01.mind0.01 \
      --flip ${analysis_dir}/penncath.geno0.01.mind0.01.merge1kgenomes-merge.missnp \
      --make-bed \
      --out ${analysis_dir}/penncath.geno0.01.mind0.01_flipSNPs
```

## Re-attempt the merge

```{bash remerge-1kg}

plink --bfile ${analysis_dir}/penncath.geno0.01.mind0.01_flipSNPs \
      --bmerge ${reference_dir}/all_pops_1kgenomes_markersForPCA.bed \
      ${reference_dir}/all_pops_1kgenomes_markersForPCA.bim \
      ${reference_dir}/all_pops_1kgenomes_markersForPCA.fam \
      --make-bed --out ${analysis_dir}/penncath.geno0.01.mind0.01.merge1kgenomes
```

## PCA on the merged dataset

The merged dataset now contains the union of PennCATH's and 1000 Genomes' variants. Most markers are genotyped in only one of the two source datasets — a marker unique to PennCATH is missing in essentially all 2,504 1000 Genomes samples (missingness ≈ 2,504/3,884 ≈ 64.5%), and a marker unique to 1000 Genomes is missing in essentially all 1,380 PennCATH samples (missingness ≈ 1,380/3,884 ≈ 35.5%). Applying `--geno 0.1` (a 10% missingness filter) drops both categories of cohort-unique markers, leaving mostly the small set of markers both cohorts genotyped in common.

(Note: the exact missingness percentage for a cohort-unique marker depends on the relative sizes of the two cohorts being merged — it isn't a fixed number in general, though in our case both categories comfortably exceed the 10% threshold.)

**Note:** unlike the study-only PCA above, we do not apply the PennCATH `.prune.in` list here because it was created from PennCATH's full marker set, not from the much smaller set of variants shared after the merge. For this teaching workflow, we use the provided `all_pops_1kgenomes_markersForPCA` marker set (already pre-pruned for LD) as supplied. In a production analysis, verify that the final shared marker set is appropriately controlled for LD before computing PCs.

```{bash merged-pca}

plink --bfile ${analysis_dir}/penncath.geno0.01.mind0.01.merge1kgenomes \
      --geno 0.1 --maf 0.05 --pca \
      --out ${analysis_dir}/penncath.geno0.01.mind0.01.merge1kgenomes
```

## Get 1KG reference ancestry labels

```{bash clone-akt}
cd ${analysis_dir}
git clone https://github.com/Illumina/akt.git
```

## Plot PCA with 1KG reference ancestry labels

```{r plot-merged-pca}
library(tidyverse)
library(data.table)
library(glue)


pc_with1kgenomes <- fread(file.path(analysis_dir, "penncath.geno0.01.mind0.01.merge1kgenomes.eigenvec"))
pop_info_1kgenomes <- fread(file.path(analysis_dir, "akt/data/integrated_call_samples_v3.20130502.ALL.panel"))

pc_with1kgenomes <- pc_with1kgenomes %>%
  rename(PC1 = V3, PC2 = V4, PC3 = V5) %>%
  left_join(pop_info_1kgenomes, by = c("V1" = "sample"))

ggplot(pc_with1kgenomes, aes(x = PC1, y = PC2)) +
  geom_point(
    data = subset(pc_with1kgenomes, !is.na(super_pop)),
    aes(color = super_pop), alpha = 0.5
  ) +
  geom_point(
    data = subset(pc_with1kgenomes, is.na(super_pop)),
    color = "black", alpha = 1
  ) +
  labs(title = "PC1 vs. PC2", color = "1KG ref ancestry labels") +
  theme_classic()

ggplot(pc_with1kgenomes, aes(x = PC2, y = PC3)) +
  geom_point(
    data = subset(pc_with1kgenomes, !is.na(super_pop)),
    aes(color = super_pop), alpha = 0.5
  ) +
  geom_point(
    data = subset(pc_with1kgenomes, is.na(super_pop)),
    color = "black", alpha = 1
  ) +
  labs(title = "PC2 vs. PC3", color = "1KG ref ancestry labels") +
  theme_classic()
```

PennCATH study samples are shown in black. In these PC plots, the PennCATH samples lie predominantly within or very close to the cluster formed by 1KG samples carrying the **EUR** reference label. Thus, at the resolution of these PCs and this reference marker set, the PennCATH samples are genetically most similar to the 1KG-defined EUR cluster. This should be interpreted as reference-based genetic similarity, not as a categorical determination that every participant is "European." Fine-scale structure, admixture, reference-panel choice, marker selection, and technical effects can all influence PC placement.
---

# QC Step 8: Genetic Ancestry Assignment with GrafAnc

In Step 7, we used joint PCA with 1KG reference-labeled samples to place PennCATH genetic variation in a reference context. It requires careful allele harmonization and several manual processing steps.

[GrafAnc](https://github.com/jimmy-penn/grafanc/blob/master/GrafAncDocumentation.md) is a purpose-built tool that estimates genetic ancestry from genotype data without requiring a study-by-study merge with an external reference panel. It uses 282,424 pre-selected biallelic ancestry SNPs to calculate reference-based genetic-distance scores and assigns continental- and subcontinental-level ancestry groups. These are algorithmic assignments relative to GrafAnc's reference populations, not direct measurements of race, ethnicity, or personal identity.

We'll run GrafAnc on our QC'd PennCATH data and compare its assignments with the reference-based PCA pattern from Step 7.

**Note:** GrafAnc requires unique sample identifiers. Because the PennCATH within-family IDs are not unique, we first create unique IIDs by combining the existing family and within-family IDs.

## Set up unique sample IDs

```{bash grafanc-uniqueids}

awk '{
    gsub("\"", "", $1);
    print "\"" $1 "\"", $2, "\"" $1 "\"", "\"" $1 "_" $2 "\""
}' ${analysis_dir}/penncath.geno0.01.mind0.01.fam > ${analysis_dir}/penncath.updateIDs.txt

plink --bfile ${analysis_dir}/penncath.geno0.01.mind0.01 --update-ids \
     ${analysis_dir}/penncath.updateIDs.txt --make-bed \
     --out ${analysis_dir}/penncath.geno0.01.mind0.01.uniqueIDs
```

## Install GrafAnc

For this workshop, we use GrafAnc directly from its GitHub repository in your analysis directory. Because that directory is persistent, the clone only needs to be created once; the code below skips the clone when it is already present.

```{bash grafanc-install}

# Clone GrafAnc (skip if already present)
if [ ! -d "${analysis_dir}/grafanc" ]; then
    cd ${analysis_dir}
    git clone https://github.com/jimmy-penn/grafanc.git
fi

# GrafAnc's reference SNP panel ships gzipped and must be decompressed once
if [ -f "${analysis_dir}/grafanc/cpp/data/AncSnpPopAFs.txt.gz" ]; then
    gunzip ${analysis_dir}/grafanc/cpp/data/AncSnpPopAFs.txt.gz
fi

ls ${analysis_dir}/grafanc/bin/
```

## Run GrafAnc

```{bash run-grafanc}
module load bcftools/1.23
export GRAFPATH=${analysis_dir}/grafanc/cpp

${analysis_dir}/grafanc/bin/grafanc \
    ${analysis_dir}/penncath.geno0.01.mind0.01.uniqueIDs \
    ${analysis_dir}/penncath_grafanc_out.txt

head ${analysis_dir}/penncath_grafanc_out.txt
```

This produces one row per sample, including continuous genetic-distance/contrast scores (`GD1-3`, `EA1-4`, `AF1-3`, `EU1-3`, `SA1-2`, `IC1-3`), three GrafAnc ancestry-component estimates (`Pe`, `Pf`, `Pa`, corresponding to European-, African-, and East-Asian-referenced components), and an assigned `AncGroupID`. The first digit of `AncGroupID` encodes the continental-level GrafAnc group (for example, `3xx` = EUR, `1xx` = AFR, `5xx` = EAS), while the full code specifies a finer reference-based group (for example, `303` = Western Europe). `Pe`, `Pf`, and `Pa` are components of GrafAnc's three-reference model; they should not be interpreted as a complete decomposition across every GrafAnc group such as South Asian or Middle East/North Africa.

## Visualize GrafAnc ancestry scores

Where does each individual's genome fall in a genetic-ancestry coordinate space defined by European, African, and East Asian reference populations?

```{r plot-grafanc}
library(tidyverse)
library(data.table)
library(glue)


grafanc_out <- fread(file.path(analysis_dir, "penncath_grafanc_out.txt"))

# GrafAnc's continental-level ancestry group is encoded in the first digit of AncGroupID
grafanc_out <- grafanc_out %>%
  mutate(
    ContinentalGroup = recode(
      AncGroupID %/% 100,
      `1` = "AFR",
      `2` = "MEN",
      `3` = "EUR",
      `4` = "SAS",
      `5` = "EAS",
      `6` = "AMR",
      `7` = "OCN",
      `8` = "MIX",
      .default = NA_character_
    )
  )

grafanc_out %>% ggplot(aes(x = GD1, y = GD2, color = ContinentalGroup)) +
  geom_point(alpha = 0.6) +
  labs(
    title = "GrafAnc-assigned genetic ancestry (GD1 vs. GD2)",
    color = "GrafAnc-assigned ancestry group"
  ) +
  theme_classic()

grafanc_out %>% count(ContinentalGroup, sort = TRUE)
```

**Note:** GD1 and GD2 are genetic-distance coordinates derived from the subject's relationship to European (E), African (F), and East Asian (A) reference populations; their position within the E–F–A reference triangle is then used to estimate continental ancestry proportions (Pe, Pf, Pa).

## Summarize GrafAnc ancestry calls for the cohort

```{r summarize-grafanc}
grafanc_out %>%
  count(AncGroupID, ContinentalGroup, sort = TRUE) %>%
  mutate(percent = round(100 * n / sum(n), 1))
```

**Question:** How do the GrafAnc assignments compare with the reference-based PCA pattern from Steps 6–7? In this dataset, you should find that >99% of samples receive GrafAnc's **EUR** continental-group assignment, distributed across several European subgroups, with a small number assigned to the **MEN** (Middle East and North Africa) group.

---

# Exercise

We now have HWE statistics for each marker, heterozygosity estimates per sample, pairwise relatedness (IBD) estimates, and GrafAnc ancestry calls for each sample. Using these:

1. Choose a HWE p-value threshold and justify your choice.
2. Choose and apply a threshold for excluding samples based on heterozygosity, and separately for relatedness. **Note:** a well-justified analysis may conclude that no samples need to be excluded on one or both of these criteria. The goal is to apply and justify a defensible method, not to guarantee you find something to remove.
3. **Optional:** Consider whether the study design or analysis plan calls for flagging samples whose genetic-structure patterns differ from the main cohort. Do not exclude a sample solely because its GrafAnc label differs from the majority. Depending on the scientific question and analysis method, defensible options can include retaining all samples with appropriate population-structure adjustment, conducting stratified analyses, or restricting an analysis to a more genetically homogeneous subset when that restriction is scientifically justified and prespecified.
4. Use PLINK to produce a QC-filtered dataset (`--hwe`, `--remove`, etc. as appropriate) based on the criteria covered in this workshop.

```{bash exercise-code}
# Your code here
```

---

# Wrap-up

Your analysis directory is on persistent Biowulf storage under your `/data/$USER` allocation, so no additional saving step is needed. Note that this now includes your `grafanc/` clone (roughly 30 MB), so you won't need to re-clone it in a future session.

To check your total disk usage for this session:

```{bash check-usage}
du -sh ${analysis_dir}
```

If you'd like to clean up any files or directories from your analysis directory to save space, you can use `rm` (for individual files) or `rm -r` (for directories) directly in a bash chunk or terminal, e.g.:

```{bash cleanup-example}
# rm ${analysis_dir}/some_file_to_delete
# rm -r ${analysis_dir}/some_directory_to_delete
```
