---
title: "Session 3: Cross-Ancestry GWAS Meta-Analysis with METASOFT"
output: html_document
---

# Objectives

- Inspect a harmonized, regional BCAC2017 dataset at the **TERT** breast-cancer susceptibility locus.
- Install and test METASOFT on Biowulf.
- Run fixed-effect, conventional random-effects, and Han-Eskin RE2 meta-analysis across AFR, ASN, and EUR ancestry-specific summary statistics.
- Use Cochran's Q, I², forest plots, and M-values to describe heterogeneity across ancestry strata.
- Visualize regional fixed-effect results, FE-versus-RE2 evidence, ancestry-specific effect estimates, and M-values.

# Prerequisites

- A Biowulf HPC account.
- RStudio launched from NIH OnDemand.

# Workshop Overview

Today we will meta-analyze AFR, ASN, and EUR ancestry-specific breast-cancer GWAS summary statistics from BCAC2017. In METASOFT terminology, each ancestry-specific GWAS stratum is treated as a separate “study” because it contributes one beta/SE pair per variant.

The raw BCAC2017 files are millions of rows, so this practical starts from a prepared **TERT** window on chromosome 5. We will focus on:

1. inspect the harmonized beta/SE input;
2. install and test METASOFT;
3. run the standard and M-value METASOFT analyses;
4. parse METASOFT's output;
5. compare fixed-effect and RE2 evidence;
6. examine heterogeneity with Q, I², ancestry-specific betas, forest plots, and M-values.

**External resources**

- Full BCAC2017 summary-statistics download: <https://drive.google.com/drive/folders/1fJbSlf0NysBZ9CQiZ3iHfU_Wb420E58V>
- METASOFT tutorial and downloads: <https://hanlab-snu.github.io/METASOFT/>

---

# Setup

## Set path variables

`input_dir` is the shared directory that contains the cleaned BCAC2017 TERT files. `analysis_dir` and `tools_dir` are writable locations for this dry run. METASOFT is small, so we install it directly inside the analysis directory.

```{r setup-paths}
username <- Sys.getenv("USER")

if (!nzchar(username)) {
  stop("USER is not set in this R session.")
}

input_dir <- file.path(
  "/data", "DCEG_shared", "statgen_workshop_2026",
  "data", "workshop2", "meta_analysis"
)
analysis_dir <- file.path(
  "/data", username, "Stats_Gen", "workshop2", "02b_analysis"
)
tools_dir <- file.path(
  analysis_dir, "metasoft"
)

dir.create(analysis_dir, recursive = TRUE, showWarnings = FALSE)
dir.create(tools_dir, recursive = TRUE, showWarnings = FALSE)

if (!dir.exists(analysis_dir)) {
  stop("Could not create analysis directory: ", analysis_dir)
}

if (!dir.exists(tools_dir)) {
  stop("Could not create tools directory: ", tools_dir)
}

Sys.setenv(
  input_dir = input_dir,
  analysis_dir = analysis_dir,
  tools_dir = tools_dir
)
```

```{bash inspect-paths}
echo "=== METASOFT input files ==="
ls -lh "${input_dir}"

echo ""
echo "=== Analysis directory ==="
ls -ld "${analysis_dir}"

echo ""
echo "=== METASOFT tools directory ==="
ls -ld "${tools_dir}"
```

## Load R libraries

```{r load-libraries}
library(data.table)
library(ggplot2)
library(glue)
```

## Define R paths

```{r define-r-paths}
data_prefix <- "bcac2017_tert_chr5_753282_1795183"

harmonized_long_file <- file.path(input_dir, glue("{data_prefix}_harmonized_long.tsv"))
variant_key_file <- file.path(input_dir, glue("{data_prefix}_variant_key.tsv"))
metasoft_input_file <- file.path(input_dir, glue("{data_prefix}_metasoft.txt"))

stopifnot(file.exists(harmonized_long_file))
stopifnot(file.exists(variant_key_file))
stopifnot(file.exists(metasoft_input_file))
```

---

# Data Overview

This practical needs three prepared files:

- `*_harmonized_long.tsv` — one row per observed `variant_id × ancestry`, used for ancestry-specific QC and forest plots.
- `*_variant_key.tsv` — one row per union variant, used to attach `chr`, `pos_b37`, `effect_allele`, and `variant_type` to METASOFT output.
- `*_metasoft.txt` — no-header METASOFT input in `AFR`, `ASN`, `EUR` order.

The shared directory also contains `*_metasoft_with_header.tsv` for debugging, but we do not need it in the analysis below. METASOFT expects a matrix with no header.

```{bash view-prepared-files}

for f in \
  ${input_dir}/bcac2017_tert_chr5_753282_1795183_harmonized_long.tsv \
  ${input_dir}/bcac2017_tert_chr5_753282_1795183_variant_key.tsv \
  ${input_dir}/bcac2017_tert_chr5_753282_1795183_metasoft.txt
do
  echo "=== ${f} ==="
  ls -lh "${f}"
  head -n 3 "${f}"
  echo ""
done
```

## Cleaning and harmonization framework

The prepared TERT dataset follows the same core logic used in meta-analysis of GWAS summary statistics:

1. **Select comparable effect estimates.** Keep one beta and standard error for the same phenotype from each ancestry-specific file.
2. **Normalize columns.** Put study-specific column names into one schema: variant ID, chromosome, position, allele pair, effect allele, beta, SE, p-value, allele frequency, and imputation quality.
3. **Use one genome build.** Here all positions are GRCh37/hg19.
4. **Resolve effect alleles.** All beta values must refer to the same allele before they can be meta-analyzed.
5. **Check variant identity.** A shared `variant_id` must map consistently to one chromosome, position, allele pair, and resolved effect allele. Variant ID alone is not sufficient for allele harmonization.
6. **Harmonize effect direction.** If an ancestry-specific result uses the opposite effect allele, multiply beta by `-1`; when EAF is defined for that same effect allele, replace EAF by `1 - EAF`.
7. **Keep a union matrix.** Variants absent from an ancestry receive `NA` for that ancestry's beta and SE.

For this class, the instructor applied those steps to the TERT region:

```text
chr5:753282-1795183, GRCh37/hg19
```

The final union contains:

```text
7,301 variants observed in 1 ancestry
2,044 variants observed in 2 ancestries
4,246 variants observed in 3 ancestries
13,591 total TERT variants
```

Palindromic SNVs and multi-base alleles were retained in this prepared dataset only after the instructor's consistency checks on variant identity and allele coding. In a new dataset, palindromic SNVs require particular care because strand orientation cannot be resolved from the allele pair alone.

## METASOFT input format

METASOFT expects one row per variant. The first column is an ID and every study contributes an adjacent beta/SE pair:

```text
variant_id  beta_AFR  se_AFR  beta_ASN  se_ASN  beta_EUR  se_EUR
```

Missing beta/SE pairs are encoded as `NA`.

---

# Step 1: Inspect the Harmonized Summary Statistics

```{r read-prepared-data}
harmonized_long <- fread(harmonized_long_file)
variant_key <- fread(variant_key_file)
metasoft_input_preview <- fread(
  metasoft_input_file,
  header = FALSE,
  col.names = c(
    "variant_id",
    "beta_AFR", "se_AFR",
    "beta_ASN", "se_ASN",
    "beta_EUR", "se_EUR"
  )
)

cat("Rows in harmonized_long:", nrow(harmonized_long), "\n")
cat("Rows in variant_key:", nrow(variant_key), "\n")
cat("Rows in METASOFT input:", nrow(metasoft_input_preview), "\n")
```

## Per-ancestry row counts

```{r ancestry-counts}
harmonized_long[, .(
  n_rows = .N,
  n_variants = uniqueN(variant_id),
  n_multi_base = sum(variant_type == "multi_base"),
  n_palindromic_snv = sum(is_palindromic_snv),
  n_effect_code_inferred = sum(effect_allele_code_inferred)
), by = ancestry][order(ancestry)]
```

## Union overlap across ancestries

A variant can be present in one, two, or all three ancestry strata. Variants present in only one ancestry remain in the long table and METASOFT input, but at least two non-missing beta/SE pairs are needed for a cross-stratum meta-analysis or heterogeneity statistic.

```{r union-overlap}
variant_overlap <- harmonized_long[, .(
  ancestries = paste(sort(unique(ancestry)), collapse = ","),
  n_ancestries = uniqueN(ancestry),
  variant_type = unique(variant_type)[1],
  is_palindromic_snv = unique(is_palindromic_snv)[1]
), by = variant_id]

variant_overlap[, .N, by = .(n_ancestries, ancestries)][order(n_ancestries, ancestries)]
```

## METASOFT input format

METASOFT expects one row per variant. The columns are:

```text
variant_id  beta_AFR  se_AFR  beta_ASN  se_ASN  beta_EUR  se_EUR
```

The no-header file starts like this:

```{r inspect-metasoft-input}
head(metasoft_input_preview)

metasoft_input_preview[, .(
  n_rows = .N,
  n_AFR = sum(!is.na(beta_AFR) & !is.na(se_AFR)),
  n_ASN = sum(!is.na(beta_ASN) & !is.na(se_ASN)),
  n_EUR = sum(!is.na(beta_EUR) & !is.na(se_EUR)),
  n_one_study = sum(rowSums(!is.na(.SD)) == 1),
  n_two_studies = sum(rowSums(!is.na(.SD)) == 2),
  n_three_studies = sum(rowSums(!is.na(.SD)) == 3)
), .SDcols = c("beta_AFR", "beta_ASN", "beta_EUR")]
```

**Question for discussion:** why should the `beta` columns for all ancestries be aligned to the same effect allele before writing this matrix?

---

# Step 2: Install METASOFT

METASOFT is not currently a Biowulf module, so for the workshop we install a copy of METASOFT into our personal `/data` directory. The package is small and contains both `Metasoft.jar` and `HanEskinPvalueTable.txt`, the tabulated p-value file used for RE2.

## Confirm Java

```{bash check-java}
which java
java -version
```

## Download and unpack METASOFT

```{bash install-metasoft}
mkdir -p "${tools_dir}"

if [[ ! -s "${tools_dir}/Metasoft.jar" || ! -s "${tools_dir}/HanEskinPvalueTable.txt" ]]; then
  module load aria2

  aria2c \
    --allow-overwrite=true \
    --dir="${tools_dir}" \
    --out=Metasoft.zip \
    "https://hanlab-snu.github.io/METASOFT/meta/repository/2.0.1/Metasoft.zip"

  unzip -o "${tools_dir}/Metasoft.zip" -d "${tools_dir}"
fi

echo "=== METASOFT files ==="
ls -lh "${tools_dir}"

test -s "${tools_dir}/Metasoft.jar"
test -s "${tools_dir}/HanEskinPvalueTable.txt"
test -s "${tools_dir}/example.txt"
```

## Confirm METASOFT can start

```{bash metasoft-help}

java -jar "${tools_dir}/Metasoft.jar" -help | head -n 25
```

## Run the bundled METASOFT example

Before analyzing the BCAC TERT data, run METASOFT's own three-study example. This separates installation problems from data-format problems.

```{bash metasoft-example}

mkdir -p "${analysis_dir}"

java -jar "${tools_dir}/Metasoft.jar" \
  -input "${tools_dir}/example.txt" \
  -output "${analysis_dir}/metasoft_example.out" \
  -log "${analysis_dir}/metasoft_example.log" \
  -pvalue_table "${tools_dir}/HanEskinPvalueTable.txt"

echo "=== Example output ==="
head "${analysis_dir}/metasoft_example.out"

echo ""
echo "=== Example log ==="
cat "${analysis_dir}/metasoft_example.log"
```

---

# Step 3: Run the Default BCAC TERT Meta-Analysis

The default METASOFT run reports fixed-effect results, conventional random-effect results, Han-Eskin RE2 results, Cochran's Q, and I².

```{bash run-metasoft-default}

mkdir -p "${analysis_dir}"

java -jar "${tools_dir}/Metasoft.jar" \
  -input "${input_dir}/bcac2017_tert_chr5_753282_1795183_metasoft.txt" \
  -output "${analysis_dir}/bcac2017_tert_metasoft.out" \
  -log "${analysis_dir}/bcac2017_tert_metasoft.log" \
  -pvalue_table "${tools_dir}/HanEskinPvalueTable.txt"

echo "=== Output rows: 13,591 variants + 1 header ==="
wc -l "${analysis_dir}/bcac2017_tert_metasoft.out"

echo ""
echo "=== First few output rows ==="
head -n 5 "${analysis_dir}/bcac2017_tert_metasoft.out"

echo ""
echo "=== METASOFT log ==="
cat "${analysis_dir}/bcac2017_tert_metasoft.log"
```

## Why does the log mention 6,290 SNPs?

The input contains 13,591 TERT variants. METASOFT keeps the one-ancestry rows in its output, but it can only compute a meta-analysis for variants observed in at least two studies.

```text
2,044 variants in exactly 2 ancestries
4,246 variants in exactly 3 ancestries
6,290 variants with at least 2 ancestries
```

So the log line `Number of SNPs analyzed: 6290` is expected.

---

# Step 4: Run METASOFT with M-Values

An M-value is METASOFT's model-based posterior probability that a nonzero effect is present in a particular study, conditional on the observed study estimates and the M-value priors.

By default, METASOFT computes M-values only for variants whose FE or RE2 p-value is below `1e-7`. For this teaching exercise, we set `-mvalue_p_thres 1` so that M-values are available across the full regional output. This is useful for visualization, but M-values should still be interpreted as posterior probabilities under METASOFT's model and priors rather than as definitive labels of whether an effect is biologically present or absent.

```{bash run-metasoft-mvalue}

java -jar "${tools_dir}/Metasoft.jar" \
  -input "${input_dir}/bcac2017_tert_chr5_753282_1795183_metasoft.txt" \
  -output "${analysis_dir}/bcac2017_tert_metasoft_mvalue.out" \
  -log "${analysis_dir}/bcac2017_tert_metasoft_mvalue.log" \
  -pvalue_table "${tools_dir}/HanEskinPvalueTable.txt" \
  -mvalue \
  -mvalue_p_thres 1

head -n 5 "${analysis_dir}/bcac2017_tert_metasoft_mvalue.out"
```

---

# Step 5: Parse the METASOFT Results

The METASOFT output is not yet convenient for plotting in R. The first 16 columns contain the meta-analysis statistics, followed by one single-study p-value and one M-value for each input study. In the output header, `PVALUES_OF_STUDIES` and `MVALUES_OF_STUDIES` are group labels rather than unique names for every study-specific column. Because our input order is AFR, ASN, EUR, we assign those names explicitly.

This parsing step is needed for three practical reasons:

1. **Give every output column a unique, meaningful name.** Otherwise the AFR/ASN/EUR p-values and M-values are difficult to address reliably in R.
2. **Check that the file structure is what we expect.** The column-count and row-count checks catch an unexpectedly formatted METASOFT output before we make plots.
3. **Restore genomic annotation.** METASOFT carries the variant ID but not our chromosome position or allele annotations, so we join the results back to `variant_key`.

## Read and name the METASOFT columns

```{r parse-metasoft-output}
metasoft_mvalue_file <- file.path(analysis_dir, "bcac2017_tert_metasoft_mvalue.out")
study_names <- c("AFR", "ASN", "EUR")

metasoft_colnames <- c(
  "variant_id", "num_study",
  "p_fe", "beta_fe", "se_fe",
  "p_re", "beta_re", "se_re",
  "p_re2", "stat1_re2", "stat2_re2",
  "p_be",
  "i_square", "q", "p_q", "tau_square",
  paste0("p_", study_names),
  paste0("m_", study_names)
)

meta_raw <- fread(
  metasoft_mvalue_file,
  header = FALSE,
  skip = 1,
  na.strings = "NA",
  fill = TRUE,
  showProgress = FALSE
)

# Some tab-delimited outputs can contain a completely empty trailing column.
is_empty_col <- function(x) {
  all(is.na(x) | trimws(as.character(x)) == "")
}

while (ncol(meta_raw) > length(metasoft_colnames) &&
       is_empty_col(meta_raw[[ncol(meta_raw)]])) {
  meta_raw[, (names(meta_raw)[ncol(meta_raw)]) := NULL]
}

if (ncol(meta_raw) != length(metasoft_colnames)) {
  stop(
    "Unexpected METASOFT column count: found ", ncol(meta_raw),
    "; expected ", length(metasoft_colnames), "."
  )
}

setnames(meta_raw, metasoft_colnames)
meta <- meta_raw
```

## Check the parsed rows and add genomic annotation

```{r annotate-metasoft-output}
meta_qc <- meta[, .N, by = num_study][order(num_study)]
print(meta_qc)

stopifnot(nrow(meta) == 13591)
stopifnot(all(meta$num_study %in% 1:3))

meta_annotated <- merge(
  meta,
  variant_key,
  by = "variant_id",
  all.x = TRUE
)

stopifnot(!anyNA(meta_annotated$pos_b37))

head(meta_annotated)
```

## Check the parsed columns

```{r inspect-parsed-output}
meta_annotated[
  ,
  .(
    n_rows = .N,
    n_finite_fe = sum(is.finite(p_fe)),
    n_finite_re2 = sum(is.finite(p_re2)),
    n_finite_q = sum(is.finite(p_q)),
    n_mvalue_AFR = sum(is.finite(m_AFR)),
    n_mvalue_ASN = sum(is.finite(m_ASN)),
    n_mvalue_EUR = sum(is.finite(m_EUR))
  ),
  by = num_study
][order(num_study)]

head(meta_annotated)
```

Rows with `num_study == 1` retain the available single-study p-value, but `p_fe`, `p_re2`, `q`, and `i_square` are `NA`: with only one non-missing ancestry-specific estimate, there is no cross-stratum meta-analysis or between-stratum heterogeneity to calculate.

---

# Step 6: Top Association and Heterogeneity Results

## Top fixed-effect hits

The fixed-effect model assumes one common underlying effect across the included ancestry strata and estimates it by inverse-variance weighting.

```{r top-fe-hits}
top_fe <- meta_annotated[
  num_study >= 2 & is.finite(p_fe)
][
  order(p_fe),
  .(
    variant_id,
    chr,
    pos_b37,
    effect_allele,
    other_allele,
    num_study,
    beta_fe,
    se_fe,
    p_fe,
    p_re2,
    i_square,
    p_q,
    variant_type
  )
]

head(top_fe, 10)
```

## Top RE2 hits

RE2 is an association test designed to retain power when effects are heterogeneous across studies. Its statistic combines evidence from a mean-effect component and a heterogeneity component. Unlike FE and the conventional random-effects model, METASOFT reports an RE2 p-value but not a single combined RE2 beta and SE.

```{r top-re2-hits}
top_re2 <- meta_annotated[
  num_study >= 2 & is.finite(p_re2)
][
  order(p_re2),
  .(
    variant_id,
    chr,
    pos_b37,
    effect_allele,
    other_allele,
    num_study,
    p_fe,
    p_re2,
    stat1_re2,
    stat2_re2,
    i_square,
    p_q
  )
]

head(top_re2, 10)
```

## High-heterogeneity three-ancestry variants

Here we only inspect variants present in all three ancestry strata. With only three estimates per variant, I² and Q are noisy measures of heterogeneity, so use this ranking as a screening device and inspect the underlying ancestry-specific estimates before drawing conclusions.

```{r top-heterogeneous-hits}
heterogeneous_hits <- meta_annotated[
  num_study == 3 &
    is.finite(p_fe) &
    is.finite(i_square)
][
  order(-i_square, p_fe),
  .(
    variant_id,
    chr,
    pos_b37,
    effect_allele,
    other_allele,
    p_fe,
    p_re2,
    i_square,
    q,
    p_q,
    p_AFR,
    p_ASN,
    p_EUR,
    m_AFR,
    m_ASN,
    m_EUR
  )
]

head(heterogeneous_hits, 10)
```

**Discussion:** compare the first rows of `top_fe`, `top_re2`, and `heterogeneous_hits`. Which variants look like shared signals, and which look ancestry-specific?

---

# Step 7: Regional and Heterogeneity Plots

## Regional fixed-effect plot

This plot is regional rather than genome-wide: we are zoomed into the TERT teaching window on chr5.

```{r plot-regional-fe, fig.width=8, fig.height=4.8}
plot_df <- meta_annotated[
  num_study >= 2 &
    is.finite(p_fe) &
    p_fe > 0
]

lead_fe <- plot_df[which.min(p_fe)]

p_regional_fe <- ggplot(
  plot_df,
  aes(x = pos_b37, y = -log10(p_fe), shape = factor(num_study))
) +
  geom_point(alpha = 0.75, size = 2) +
  geom_point(
    data = lead_fe,
    aes(
      x = pos_b37,
      y = -log10(p_fe),
      shape = factor(num_study)
    ),
    inherit.aes = FALSE,
    size = 4
  ) +
  annotate(
    "text",
    x = lead_fe$pos_b37,
    y = -log10(lead_fe$p_fe),
    label = lead_fe$variant_id,
    hjust = -0.05,
    vjust = -0.4,
    size = 3
  ) +
  scale_shape_manual(
    values = c("2" = 17, "3" = 16),
    labels = c("2" = "2", "3" = "3")
  ) +
  scale_x_continuous(labels = scales::comma) +
  scale_y_continuous(
    limits = c(0, NA),
    expand = expansion(mult = c(0.02, 0.10))
  ) +
  labs(
    x = "GRCh37 position on chromosome 5",
    y = expression(-log[10](P[FE])),
    shape = "No. of ancestry strata",
    title = "TERT regional association: fixed-effect meta-analysis"
  ) +
  theme_bw()

p_regional_fe
```

## Fixed-effect versus RE2 p-values

This plot compares the strength of association evidence from FE and RE2 for the same variants. Points above the dashed line have smaller p-values under RE2 than under FE; color shows I² to help identify whether departures from the diagonal coincide with estimated heterogeneity.

```{r plot-fe-vs-re2, fig.width=6, fig.height=5}
fe_re2_df <- meta_annotated[
  num_study >= 2 &
    is.finite(p_fe) & p_fe > 0 &
    is.finite(p_re2) & p_re2 > 0
]

max_logp <- max(-log10(fe_re2_df$p_fe), -log10(fe_re2_df$p_re2), na.rm = TRUE)

ggplot(
  fe_re2_df,
  aes(x = -log10(p_fe), y = -log10(p_re2), color = i_square)
) +
  geom_abline(slope = 1, intercept = 0, linetype = "dashed") +
  geom_point(alpha = 0.8, size = 2) +
  coord_equal(xlim = c(0, max_logp), ylim = c(0, max_logp)) +
  labs(
    x = expression(-log[10](P[FE])),
    y = expression(-log[10](P[RE2])),
    color = expression(I^2),
    title = "Fixed-effect versus RE2 evidence"
  ) +
  theme_bw()
```

---

# Step 8: Forest Plots for Individual Variants

A forest plot shows the ancestry-specific log odds-ratio estimates and 95% confidence intervals that contribute to the meta-analysis.

## Fixed-effect lead variant

```{r plot-lead-fe-forest, fig.width=7, fig.height=3.5}
lead_variant <- top_fe$variant_id[1]

ancestry_df <- harmonized_long[
  variant_id == lead_variant,
  .(
    group = ancestry,
    beta,
    se,
    estimate_type = "Ancestry-specific"
  )
]

meta_df <- meta_annotated[
  variant_id == lead_variant,
  .(
    group = "Fixed effect",
    beta = beta_fe,
    se = se_fe,
    estimate_type = "Meta-analysis"
  )
]

forest_df <- rbindlist(
  list(ancestry_df, meta_df),
  use.names = TRUE
)

forest_df[
  ,
  `:=`(
    group = factor(group, levels = c("Fixed effect", "EUR", "ASN", "AFR")),
    ci_low = beta - 1.96 * se,
    ci_high = beta + 1.96 * se
  )
]

ggplot(
  forest_df,
  aes(
    y = group,
    x = beta,
    xmin = ci_low,
    xmax = ci_high,
    shape = estimate_type
  )
) +
  geom_vline(xintercept = 0, linetype = "dashed") +
  geom_pointrange(size = 0.7) +
  labs(
    x = "log odds ratio per effect allele",
    y = NULL,
    shape = NULL,
    title = paste("Ancestry-specific and fixed-effect estimates for", lead_variant)
  ) +
  theme_bw()
```

The fixed-effect estimate is inverse-variance weighted, so estimates with smaller SEs receive more weight.


## A high-I² variant

```{r plot-high-i2-forest, fig.width=7, fig.height=3.5}
high_i2_variant <- heterogeneous_hits$variant_id[1]

high_i2_ancestry_df <- harmonized_long[
  variant_id == high_i2_variant,
  .(
    estimate = ancestry,
    beta,
    se,
    estimate_type = "Ancestry-specific"
  )
]

high_i2_meta_df <- meta_annotated[
  variant_id == high_i2_variant,
  .(
    estimate = "Fixed effect",
    beta = beta_fe,
    se = se_fe,
    p_fe,
    p_re2,
    i_square,
    p_q,
    m_AFR,
    m_ASN,
    m_EUR,
    estimate_type = "Meta-analysis"
  )
]

high_i2_forest_df <- rbindlist(
  list(
    high_i2_ancestry_df[
      ,
      .(
        estimate,
        beta,
        se,
        p_fe = NA_real_,
        p_re2 = NA_real_,
        i_square = NA_real_,
        p_q = NA_real_,
        m_AFR = NA_real_,
        m_ASN = NA_real_,
        m_EUR = NA_real_,
        estimate_type
      )
    ],
    high_i2_meta_df
  ),
  use.names = TRUE
)

high_i2_forest_df[
  ,
  `:=`(
    estimate = factor(estimate, levels = c("Fixed effect", "EUR", "ASN", "AFR")),
    ci_low = beta - 1.96 * se,
    ci_high = beta + 1.96 * se
  )
]

ggplot(
  high_i2_forest_df,
  aes(
    y = estimate,
    x = beta,
    xmin = ci_low,
    xmax = ci_high,
    shape = estimate_type
  )
) +
  geom_vline(xintercept = 0, linetype = "dashed") +
  geom_pointrange(size = 0.8) +
  labs(
    x = "log odds ratio per effect allele",
    y = NULL,
    shape = NULL,
    title = paste("Ancestry-specific and fixed-effect estimates for", high_i2_variant),
    subtitle = paste0(
      "I² = ", round(high_i2_meta_df$i_square, 1),
      "%; FE p = ", formatC(high_i2_meta_df$p_fe, format = "e", digits = 2),
      "; RE2 p = ", formatC(high_i2_meta_df$p_re2, format = "e", digits = 2),
      "; M(AFR, ASN, EUR) = ",
      paste(round(c(high_i2_meta_df$m_AFR, high_i2_meta_df$m_ASN, high_i2_meta_df$m_EUR), 2),
            collapse = ", ")
    )
  ) +
  theme_bw()
```

For this variant, compare the fixed-effect estimate with the ancestry-specific estimates rather than assuming that a high I² has a single cause. RE2 does not report its own combined beta; its p-value uses both mean-effect and heterogeneity components. The M-values provide model-based evidence about which ancestry strata are more compatible with a nonzero effect, but they do not by themselves establish ancestry-specific biological effect modification.

---

# Step 9: M-Value Heatmap

The M-value columns give METASOFT's posterior probability, under its specified priors and model, that a nonzero effect is present in each ancestry-specific study. Values near 1 support effect presence, values near 0 support effect absence, and intermediate values are ambiguous.

```{r plot-mvalue-heatmap, fig.width=7, fig.height=6}
mvalue_top <- head(
  meta_annotated[num_study == 3 & is.finite(p_re2)][order(p_re2)],
  20
)

mvalue_plot_df <- melt(
  mvalue_top[, .(variant_id, m_AFR, m_ASN, m_EUR)],
  id.vars = "variant_id",
  variable.name = "ancestry",
  value.name = "mvalue"
)

mvalue_plot_df[, ancestry := sub("^m_", "", ancestry)]
mvalue_plot_df[, variant_id := factor(variant_id, levels = rev(mvalue_top$variant_id))]

ggplot(mvalue_plot_df, aes(x = ancestry, y = variant_id, fill = mvalue)) +
  geom_tile() +
  labs(
    x = NULL,
    y = NULL,
    fill = "M-value",
    title = "Posterior probability of an effect by ancestry",
    subtitle = "Top 20 three-ancestry variants ranked by RE2 p-value"
  ) +
  theme_bw()
```

# Wrap-up

In this practical, we used one harmonized regional summary-statistics file to run three complementary analyses:

- **Fixed-effect meta-analysis**: estimates one common effect across the included strata using inverse-variance weighting.
- **RE2**: tests for association using both mean-effect and heterogeneity information and can gain power when effects differ across studies.
- **M-values**: model-based posterior probabilities that help describe which studies are more compatible with effect presence.

In this workshop, differences across ancestry strata are useful for learning about heterogeneity, but they should not automatically be interpreted as ancestry-caused biological differences. Sampling variation, linkage disequilibrium, allele frequency, imputation quality, and study-design differences can also contribute to cross-stratum differences.

```{bash check-usage}

echo "=== Analysis directory ==="
du -sh "${analysis_dir}"

echo "=== METASOFT tools directory ==="
du -sh "${tools_dir}"
```
