---
title: "Session 6: Polygenic Risk Scores"
subtitle: "Clumping and thresholding in RStudio on Biowulf"
output:
  html_document:
    toc: true
    toc_depth: 2
    number_sections: true
    code_folding: show
    self_contained: true
    theme: readable
params:
  execute: true
  data_dir: "/data/DCEG_shared/statgen_workshop_2026/data/workshop6/data/"
  output_dir: "06_PRS_results"
  threads: 1
  memory_mb: 2048
  clump_r2: 0.1
  clump_kb: 500
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE, error = FALSE, warning = TRUE,
                     message = FALSE, cache = FALSE, eval = params$execute,
                     fig.width = 7, fig.height = 4.5)
```

# Overview

This practical introduces polygenic risk scores using chromosome 22 GWAS
summary statistics, a European 1000 Genomes LD reference, simulated target
genotypes, and a quantitative outcome.

We will:

1. Use PLINK 1.9 to select approximately independent variants by LD clumping.
2. Use PLINK 2 to calculate a PRS at nine GWAS p-value thresholds.
3. Select a threshold using 10,000 tuning samples.
4. Evaluate the selected score in 10,000 separate validation samples.

For individual \(i\), the score is
\[
\mathrm{PRS}_i = \sum_{j \in S} \widehat\beta_j G_{ij},
\]
where \(S\) is the selected variant set, \(\widehat\beta_j\) is the GWAS
coefficient, and \(G_{ij}\) counts copies of the **effect allele A1**.
This example predicts a quantitative trait, so performance is measured with
linear-regression \(R^2\).

# Prepare RStudio on Biowulf

Start **RStudio Server through Biowulf OnDemand** on a compute node. Start with
**2 CPUs, 8 GB RAM, and 1 hour** for this chromosome 22 exercise; request 5 GB
of local scratch if the launch form allows it. The code defaults to one PLINK
thread and a 2,048 MB PLINK workspace. Keep these settings within your allocation.

## Get your own copy

In the RStudio **Terminal** tab, clone the workshop into a directory you own:

```bash
mkdir -p "/data/$USER/Stats_Gen/workshop6"
cd "/data/$USER/Stats_Gen/workshop6"
git clone https://github.com/DCEG-workshops/statgen_workshop_tutorial.git
cd statgen_workshop_tutorial/src
pwd
```

If you already have a clone, use that copy. In RStudio, select **File > Open
File...** and open `06_PRS.Rmd` in the `src` directory printed above. Select
**Session > Set Working Directory > To Source File Location** before running
individual chunks. The Terminal's `cd` command does not change R's working
directory. Run the chunks in order, or click **Knit** to generate an HTML report.

The R packages `data.table`, `knitr`, and `rmarkdown` must be available. Biowulf's
`R/4.5` module provides them. This document loads `plink/1.9.0-beta4.4` for
clumping and `plink/7.1-alpha` (PLINK 2) for scoring in separate shell processes.

Alternatively, from the `src` directory in a **compute-node terminal**:

```bash
module load R/4.5
Rscript -e 'rmarkdown::render("06_PRS.Rmd")'
```

To create a code-only preview without running the analyses:

```r
rmarkdown::render("06_PRS.Rmd", params = list(execute = FALSE),
                  output_file = "06_PRS.preview.html")
```

```{r preview-label, eval=TRUE, echo=FALSE, results='asis'}
if (!isTRUE(params$execute)) {
  cat("**Code-only preview: no analyses or data checks were run.**\n\n")
}
```

## Input files and output directory

The `data_dir` parameter points directly to the supplied `data/` folder. Only
the following files are required:

| Path relative to `data_dir` | Role |
|:---|:---|
| `EUR_sum_data` | GWAS statistics: `SNP`, `A1`, `BETA`, `P` |
| `1kg_eur_22/chr_22.bed`, `.bim`, `.fam` | European LD reference |
| `prs_genotype/chr22_test.bed`, `.bim`, `.fam` | Target genotypes |
| `y_out` | Quantitative outcomes with columns `ID` and `y` |

Each run creates a new subdirectory of `06_PRS_results/` beside your copy of
the tutorial. The shared inputs are read only. Change `output_dir` in the YAML
header to use another writable location. Logs, scores, coefficients, the
sample split, and performance tables are saved there.

```{r paths-and-resources}
library(data.table)
setDTthreads(1)
if (!nzchar(Sys.getenv("SLURM_JOB_ID"))) {
  stop("Run this tutorial inside a Biowulf compute allocation.")
}
threads <- as.integer(params$threads)
allocated_cpus <- as.integer(Sys.getenv("SLURM_CPUS_PER_TASK", "1"))
if (is.na(threads) || threads < 1L || threads > allocated_cpus) {
  stop("threads must be between 1 and SLURM_CPUS_PER_TASK (or 1 if unset).")
}
memory_mb <- as.integer(params$memory_mb)
if (is.na(memory_mb) || memory_mb < 640L) stop("memory_mb must be at least 640.")
if (!is.finite(params$clump_r2) || params$clump_r2 <= 0 || params$clump_r2 >= 1 ||
    !is.finite(params$clump_kb) || params$clump_kb <= 0) {
  stop("Use 0 < clump_r2 < 1 and clump_kb > 0.")
}
Sys.setenv(OMP_NUM_THREADS = "1", OPENBLAS_NUM_THREADS = "1", MKL_NUM_THREADS = "1")

data_dir <- normalizePath(params$data_dir, mustWork = TRUE)
sum_data_file <- file.path(data_dir, "EUR_sum_data")
ref_data <- file.path(data_dir, "1kg_eur_22", "chr_22")
geno_file <- file.path(data_dir, "prs_genotype", "chr22_test")
outcome_file <- file.path(data_dir, "y_out")
inputs <- c(sum_data_file, paste0(ref_data, c(".bed", ".bim", ".fam")),
            paste0(geno_file, c(".bed", ".bim", ".fam")), outcome_file)
missing_inputs <- inputs[!file.exists(inputs) | file.access(inputs, 4) != 0]
if (length(missing_inputs)) stop("Missing/unreadable input: ",
                                 paste(missing_inputs, collapse = ", "))

output_dir <- normalizePath(path.expand(params$output_dir), mustWork = FALSE)
if (output_dir == data_dir || startsWith(output_dir, paste0(data_dir, "/"))) {
  stop("Choose an output_dir outside the shared input directory.")
}
dir.create(output_dir, recursive = TRUE, showWarnings = FALSE)
run_dir <- tempfile(pattern = paste0("run_", format(Sys.time(), "%Y%m%d_%H%M%S"), "_"),
                    tmpdir = output_dir)
if (!dir.create(run_dir)) stop("Cannot create a results directory: ", run_dir)
run_dir <- normalizePath(run_dir)
saveRDS(params, file.path(run_dir, "parameters.rds"))
writeLines(c(paste("Host:", Sys.info()[["nodename"]]),
             paste("Job:", Sys.getenv("SLURM_JOB_ID")),
             paste("Working directory:", getwd()),
             paste("Started:", Sys.time())), file.path(run_dir, "run_info.txt"))
knitr::kable(data.frame(input = sub(paste0(data_dir, "/"), "", inputs, fixed = TRUE),
                        MB = round(file.info(inputs)$size / 1024^2, 2)))
cat("Results directory:", run_dir, "\n")
```

## Run PLINK from R

This helper loads the appropriate Biowulf module in a child shell, records the
command and its output, and stops if PLINK fails. Switching PLINK versions
does not change the RStudio environment.

```{r plink-helper}
run_plink <- function(version, args, label) {
  module_name <- switch(version, "1" = "plink/1.9.0-beta4.4", "2" = "plink/7.1-alpha")
  binary <- switch(version, "1" = "plink", "2" = "plink2")
  if (is.null(module_name)) stop("Unknown PLINK version.")
  command <- paste(c(binary, shQuote(as.character(args))), collapse = " ")
  script <- file.path(run_dir, paste0(label, ".sh"))
  log <- file.path(run_dir, paste0(label, ".console.log"))
  writeLines(c("#!/bin/bash", "set -eo pipefail",
               "if ! type module >/dev/null 2>&1; then source /etc/profile.d/modules.sh; fi",
               paste("module load", shQuote(module_name)), command), script)
  status <- system2("bash", shQuote(script), stdout = log, stderr = log)
  if (status != 0L) stop("PLINK failed; inspect ", log)
  invisible(log)
}
version_logs <- c(run_plink("1", "--version", "plink1_version"),
                  run_plink("2", "--version", "plink2_version"))
cat(unlist(lapply(version_logs, readLines)), sep = "\n")
resource_args <- c("--threads", threads, "--memory", memory_mb)
```

# Inspect the summary statistics and outcomes

`BETA` is the effect per copy of `A1`; it is not an odds ratio. The supplied
files are prepared with consistent variant identifiers and allele coding.
For another dataset, verify genome build, strand, both alleles, and independence
of discovery and target samples before scoring. Matching an rsID alone is
insufficient to establish allele alignment.

```{r read-data}
EUR_sum <- fread(sum_data_file, colClasses = c(SNP = "character", A1 = "character"))
required_columns <- c("SNP", "A1", "BETA", "P")
if (!all(required_columns %in% names(EUR_sum))) stop("Summary-statistic columns are missing.")
if (anyNA(EUR_sum[, ..required_columns]) || anyDuplicated(EUR_sum$SNP) ||
    any(!nzchar(EUR_sum$SNP)) || any(!nzchar(EUR_sum$A1)) ||
    any(!is.finite(EUR_sum$BETA)) || any(!is.finite(EUR_sum$P)) ||
    any(EUR_sum$P < 0 | EUR_sum$P > 1)) stop("Invalid summary statistics.")

fam <- fread(paste0(geno_file, ".fam"), header = FALSE, colClasses = "character")
setnames(fam, c("FID", "IID", "father", "mother", "sex", "phenotype"))
y_out <- fread(outcome_file, colClasses = c(ID = "character"))
if (!all(c("ID", "y") %in% names(y_out))) stop("y_out must contain ID and y.")
if (nrow(y_out) != 20000L || nrow(fam) != 20000L ||
    anyNA(y_out$ID) || anyNA(fam$IID) || anyDuplicated(y_out$ID) ||
    anyDuplicated(fam$IID) || !setequal(y_out$ID, fam$IID) ||
    any(!is.finite(y_out$y))) {
  stop("Expected 20,000 finite outcomes and uniquely matching target IIDs.")
}
# Preserve phenotype order for the split; attach genotype identifiers explicitly.
samples <- data.table(FID = fam$FID[match(y_out$ID, fam$IID)],
                      IID = y_out$ID, y = y_out$y,
                      cohort = rep(c("tuning", "validation"), each = 10000L))
fwrite(samples[, .(FID, IID, cohort)], file.path(run_dir, "sample_split.tsv"), sep = "\t")
knitr::kable(data.frame(summary_variants = nrow(EUR_sum), target_samples = nrow(samples),
                        tuning_samples = sum(samples$cohort == "tuning"),
                        validation_samples = sum(samples$cohort == "validation")))
```

The first 10,000 rows of `y_out` form the tuning set and the last 10,000 form
the validation set. In the supplied data these are IDs 100001–110000 and
110001–120000, respectively.
Do not use validation outcomes to choose the threshold.

# LD clumping with PLINK 1.9

Clumping selects a lead SNP and groups nearby SNPs correlated with it, starting
with the smallest GWAS p-value. We use a **500 kb radius**, **LD \(r^2 = 0.1\)**,
and `--clump-p1 1` so index SNPs across the p-value range can be considered.
LD \(r^2\) measures correlation between variants and differs from
the prediction \(R^2\) calculated later.

```{r clump}
clump_prefix <- file.path(run_dir, "LD_clump")
run_plink("1", c("--bfile", ref_data, "--clump", sum_data_file,
                 "--clump-p1", "1", "--clump-r2", params$clump_r2,
                 "--clump-kb", params$clump_kb,
                 resource_args, "--out", clump_prefix), "clumping")
clump_file <- paste0(clump_prefix, ".clumped")
if (!file.exists(clump_file)) stop("No clumps were produced; inspect the PLINK log.")
LD_clump <- fread(clump_file)
if (!"SNP" %in% names(LD_clump) || !nrow(LD_clump)) stop("No lead SNPs were found.")
clumped <- EUR_sum[match(LD_clump$SNP, EUR_sum$SNP), .(SNP, A1, BETA, P)]
if (anyNA(clumped)) stop("Clumped SNPs did not match the summary statistics.")

# Clump all summary variants, then retain scoreable target variants.
target_bim <- fread(paste0(geno_file, ".bim"), header = FALSE, colClasses = "character")
setnames(target_bim, c("CHR", "SNP", "CM", "BP", "allele1", "allele2"))
if (anyDuplicated(target_bim$SNP)) stop("Duplicate target variant IDs.")
available <- clumped$SNP %in% target_bim$SNP
cat("LD-clumped lead SNPs:", nrow(clumped),
    "\nLead SNPs absent from target:", sum(!available), "\n")
prs_prep <- clumped[available]
target_index <- match(prs_prep$SNP, target_bim$SNP)
allele_match <- prs_prep$A1 == target_bim$allele1[target_index] |
               prs_prep$A1 == target_bim$allele2[target_index]
if (!nrow(prs_prep) || any(!allele_match)) {
  stop("No usable variants, or effect-allele mismatch; check input harmonization.")
}
setorder(prs_prep, P, SNP)
fwrite(prs_prep, file.path(run_dir, "clumped_score_candidates.tsv"), sep = "\t")
knitr::kable(head(prs_prep)[, .(SNP, A1, BETA, P = format(P, scientific = TRUE))])
```

# Calculate PRS at nine p-value thresholds

For each threshold, retain clumped SNPs with `P <= threshold` and write their
`SNP`, `A1`, and `BETA` values. `header` skips the coefficient-file header;
`1 2 3` identifies the input columns. `SCORE1_SUM` is the weighted allele sum.
With `no-mean-imputation`, a missing genotype contributes no term to the sum. Missingness can affect scores in real data.

The reference has many variants absent from the target. Clumping uses the full
summary statistics. Lead SNPs absent from the target are omitted before scoring. The table reports
both the selected lead SNPs and the SNPs actually scored at each threshold.

```{r score-thresholds}
pthres <- c(5e-8, 5e-7, 5e-6, 5e-5, 5e-4, 5e-3, 5e-2, 5e-1, 1)
performance <- data.table(threshold = pthres, n_lead_snps = 0L, n_scored_snps = 0L,
                           tuning_R2 = NA_real_, status = "not scored")
for (k in seq_along(pthres)) {
  prs_coeff <- prs_prep[P <= pthres[k], .(SNP, A1, BETA)]
  performance$n_lead_snps[k] <- sum(clumped$P <= pthres[k])
  performance$n_scored_snps[k] <- nrow(prs_coeff)
  if (!nrow(prs_coeff)) {
    performance$status[k] <- "no SNPs at this threshold"
    next
  }
  coefficient_file <- file.path(run_dir, paste0("prs_coeff_", k, ".tsv"))
  score_prefix <- file.path(run_dir, paste0("prs_", k))
  fwrite(prs_coeff, coefficient_file, sep = "\t")
  run_plink("2", c("--bfile", geno_file,
                   "--score", coefficient_file, "1", "2", "3", "header",
                   "no-mean-imputation", "list-variants", "cols=+fid,+scoresums,-scoreavgs",
                   resource_args, "--out", score_prefix), paste0("scoring_", k))
  used_snps <- readLines(paste0(score_prefix, ".sscore.vars"))
  if (!setequal(used_snps, prs_coeff$SNP)) stop("PLINK skipped scoring variants; inspect its log.")
  performance$status[k] <- "scored"
}
knitr::kable(performance[, .(threshold = format(threshold, scientific = TRUE),
                              n_lead_snps, n_scored_snps, status)])
```

# Select the threshold using the tuning set

Read scores by **FID and IID**, and reorder them to match the outcome table.
An empty SNP set or a constant score cannot be evaluated and is excluded from
threshold selection. Validation outcomes are not used here.

```{r tune}
read_scores <- function(k) {
  scores <- fread(file.path(run_dir, paste0("prs_", k, ".sscore")),
                   colClasses = "character")
  setnames(scores, sub("^#", "", names(scores)))
  if (!all(c("FID", "IID", "SCORE1_SUM") %in% names(scores))) {
    stop("Missing expected columns in PLINK score output.")
  }
  key <- paste(scores$FID, scores$IID, sep = "\t")
  expected_key <- paste(samples$FID, samples$IID, sep = "\t")
  if (anyDuplicated(key) || !setequal(key, expected_key)) {
    stop("Scored sample IDs do not match the outcomes.")
  }
  score <- as.numeric(scores$SCORE1_SUM[match(expected_key, key)])
  if (any(!is.finite(score))) stop("Non-finite PRS values.")
  score
}
tuning <- samples$cohort == "tuning"
if (var(samples$y[tuning]) == 0) stop("The tuning outcome is constant.")
for (k in which(performance$status == "scored")) {
  score <- read_scores(k)
  if (var(score[tuning]) == 0) {
    performance$status[k] <- "constant tuning score"
    next
  }
  model <- lm(y ~ PRS, data = data.frame(y = samples$y[tuning], PRS = score[tuning]))
  performance$tuning_R2[k] <- summary(model)$r.squared
}
eligible <- which(is.finite(performance$tuning_R2))
if (!length(eligible)) stop("No threshold produced an evaluable tuning score.")
# Thresholds are ascending; a tie selects the smaller threshold.
best_k <- eligible[which.max(performance$tuning_R2[eligible])]
fwrite(performance, file.path(run_dir, "tuning_performance.tsv"), sep = "\t")
performance_display <- copy(performance)
performance_display[, threshold := format(threshold, scientific = TRUE)]
knitr::kable(performance_display, digits = 6)
plot(performance$threshold, performance$tuning_R2, log = "x", type = "b", pch = 16,
     xlab = "GWAS p-value threshold", ylab = expression("Tuning " * R^2),
     main = "Choose the threshold using the tuning set")
points(performance$threshold[best_k], performance$tuning_R2[best_k],
       pch = 19, col = "firebrick", cex = 1.5)
```

# Evaluate the selected score in the validation set

Evaluate **only the selected threshold**. Fit `y ~ PRS` in the validation set
and report its \(R^2\). For one score and
an intercept, this equals squared Pearson correlation. This measures PRS
association in held-out samples; it does not test calibration of a prediction
equation frozen in the tuning set.

```{r validate}
validation <- samples$cohort == "validation"
best_score <- read_scores(best_k)
if (var(samples$y[validation]) == 0 || var(best_score[validation]) == 0) {
  stop("Validation R-squared is undefined for a constant outcome or score.")
}
validation_model <- lm(y ~ PRS, data = data.frame(
  y = samples$y[validation], PRS = best_score[validation]))
validation_result <- data.table(
  selected_threshold = pthres[best_k], n_scored_snps = performance$n_scored_snps[best_k],
  tuning_n = sum(tuning), tuning_R2 = performance$tuning_R2[best_k],
  validation_n = sum(validation), validation_R2 = summary(validation_model)$r.squared)
fwrite(validation_result, file.path(run_dir, "validation_performance.tsv"), sep = "\t")
validation_display <- copy(validation_result)
validation_display[, selected_threshold := format(selected_threshold, scientific = TRUE)]
knitr::kable(validation_display, digits = 6)
saveRDS(validation_model, file.path(run_dir, "validation_model.rds"))
```

## Questions

1. How does the number of SNPs change as the p-value threshold increases?
2. Does the threshold with the most SNPs have the highest tuning \(R^2\)?
3. Why would choosing a threshold using validation outcomes give an optimistic
   performance estimate?
4. Why might a PRS using European GWAS and LD reference data perform differently
   in another ancestry group?
5. What additional checks and covariates would an observational study need?

This unadjusted chromosome 22 simulation illustrates the workflow. It is not a
genome-wide prediction model or an estimate of clinical performance.

# Record the software environment

```{r session-info}
writeLines(capture.output(sessionInfo()), file.path(run_dir, "sessionInfo.txt"))
sessionInfo()
cat("Saved results:", run_dir, "\n")
```

Results are from the current run. PLINK `.log` files and `.console.log` files
contain variant counts, warnings, and versions; `.sh` files record the commands.

Documentation: [PLINK 1.9 clumping](https://www.cog-genomics.org/plink/1.9/postproc#clump),
[PLINK 2 scoring](https://www.cog-genomics.org/plink/2.0/score), and
[Biowulf PLINK modules](https://hpc.nih.gov/apps/plink.html).
